---
title: "Data visualization: Design choices"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data visualization: Design choices"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio, 
  skimr
)

theme_set(theme_grey(base_size = 24)) 
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data) |>
  relocate(diab, degree, .after = id)
```

## Recall: common components of `ggplot2`

::: columns
::: column
- Common components of `ggplot2` include: 
  - `ggplot()`: the main function to create a plot
  - `aes()`: the function to specify the mapping of variables to axes
  - `geom_*()` or `stat_*()`: functions to add layers to the plot
  - `labs()`: the function to change labels of axes and title
  - `theme()`: the function to adjust non-data plot elements such as text size,
:::
::: column

 

```{r}
#| eval: false

ggplot(
  data = my_data, 
  mapping = aes(x = col1, y = col2)
  ) +  #<1>       
  
  geom_point() +          
  
  labs() +       
  
  theme()                                  
```
:::
:::

::: blue
::: blue-cont
In this lesson, we will focus on **using different geometric objects** and **building the mapping**.
:::
:::

# Geometric objects

## Geometric objects

- Geometric objects (geoms) are used to represent the datapoints
  - The same variable's data can be represented in different ways

For example: we can plot Age vs. Income using a scatterplot or a smoothed line

::: columns
::: column
```{r}
#| fig-width: 10
#| fig-height: 7
#| fig-align: center
#| echo: false

ggplot(
  hrs_data, 
  mapping = aes(
    x = age_yr, 
    y = income
    )
  ) + 
  geom_point()
```
:::
::: column
```{r}
#| fig-width: 10
#| fig-height: 7
#| fig-align: center
#| echo: false

ggplot(
  hrs_data, 
  mapping = aes(
    x = age_yr, 
    y = income
    )
  ) + 
  geom_smooth()
```
:::
:::

## Types of geoms

- Common geoms include:
  - `geom_histogram()`
  - `geom_bar()`
  - `geom_boxplot()`
  - `geom_point()`
  - `geom_line()`
  - `geom_smooth()`

## Let's check out the `ggplot2` cheatsheet together

## EXAMPLES

# Mapping

## Aesthetic mapping 

- If we want to change the visual properties of the plotted data, we use **aesthetic mapping**
  - This is not about the design choices of the figure, but the aesthetics of the **data**

- Here are a list of common aesthetic options (from the [Epi Handbook](https://epirhandbook.com/en/new_pages/ggplot_basics.html#plot-aesthetics)):
  - `shape` = Display a point with geom_point() as a dot, star, triangle, or square…
  - `fill` = The interior color (e.g. of a bar or boxplot)
  - `color` = The exterior line of a bar, boxplot, etc., or the point color if using geom_point()
  - `size` = Size (e.g. line thickness, point size)
  - `alpha` = Transparency (1 = opaque, 0 = invisible)
  - `binwidth` = Width of histogram bins (for `geom_histogram()` only)
  - `width` = Width of bar plot columns (for `geom_bar()` only)
  - `linetype` = Line type (e.g. solid, dashed, dotted)
  
## Resources

- [`ggplot2` Cheatsheet in html form](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
- [`ggplot2` Cheatsheet in pdf form](https://rstudio.github.io/cheatsheets/data-visualization.pdf)



