---
title: "Key operators and functions"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Key operators and functions"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  janitor,
  here, 
  rio
)
```

```{r}
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_data <- clean_names(hrs_data_rds)
```


## Note

- This lesson is **heavily** borrowed from [Section 8.2 in The Epidemiologist R Handbook](https://epirhandbook.com/en/new_pages/basics.html#operators)

## Let's expand on some basic operators

We went over some operators in our lesson on Basics

::::: columns
::: {.column width="50%"}
Basic math operators:

| Operation      | Written                   | In R                   |
|----------------|---------------------------|------------------------|
| Addition       | $1+1$                     | `1+1`                  |
| Subtraction    | $2-1$                     | `2-1`                  |
| Multiplication | $3\times 2$ or $3\cdot 2$ | `3*2`                  |
| Division       | $3 / 2$ or $\dfrac{3}{2}$ | `3/2`                  |
| Exponential    | $4^2$                     | `4^2`                  |
| Square root    | $\sqrt{4}$ or $4^{1/2}$   | `4^(1/2)` or `sqrt(4)` |
:::

::: {.column width="5%"}
:::

::: {.column width="45%"}
Basic relational operators:

| Operation                | In R   |
|--------------------------|--------|
| Less than                | `1<1`  |
| Greater than             | `1>1`  |
| Less than or equal to    | `1<=1` |
| Greater than or equal to | `1>=1` |
| Equal to                 | `1==1` |
| Not equal to             | `1!=1` |
:::
:::::

## We can further expand the relational operators

- If we want combine two logical vectors, we need to use boolean algebra

+----------------+----------+----------------------+---------------------------------------------------------------------------------+
| Operation      | In R     | Example in R         | Example in plain language                                                       |
+================+==========+======================+=================================================================================+
| Group criteria | `( )`    | `(1 == 1)`           | We are grouping this statement to make it easier to read                        |
+----------------+----------+----------------------+---------------------------------------------------------------------------------+
| And            | `&`      | `(1 == 1) & (3 > 2)` | Is 1 equal to 1 AND 3 greater than 2? (If BOTH are true, returns true)          |
+----------------+----------+----------------------+---------------------------------------------------------------------------------+
| Or             | `|`      | `(1 == 1) | (1 > 2)` | Is 1 equal to 1 OR is 1 greater than 2? (If one of them are true, returns TRUE) |
+----------------+----------+----------------------+---------------------------------------------------------------------------------+
| Not            | `!`      | `!(1 > 2)`           | Is 1 NOT greater than 2?                                                        |
+----------------+----------+----------------------+---------------------------------------------------------------------------------+

![Adapted from this [R for Data Science](https://r4ds.hadley.nz/logicals.html#fig-bool-ops)](../img_slides/boolean.png)

## Is something in something else?

- The `%in%` operator is helpful when you want to check if one thing is **in a list of other things**

- Let's say I have a vector

```{r}
my_vector <- c("a", "b", "c", "d") #<1>
```

1.  We create a vector with letters a-d

- I want to ask: "Is "a" in my vector?"

```{r}
"a" %in% my_vector #<2>
```

2.  I want to see if `"a"` is in the vector

```{r}
"m" %in% my_vector #<3>
```

3.  Then I want to see if `"n"` is in the vector

```{r}
"m" %in% my_vector | "a" %in% my_vector #<4>
```

4.  We can combine operators as well! I'm asking: "Is "m" in my vector or "a" in my vector?"

```{r}
"A" %in% my_vector #<5>
```

5.  It is also case sensitive

## Some good functions to know

+--------------------+-----------------------------------------+-------------------------------------------+
| Purpose            | Function                                | Example                                   |
+====================+=========================================+===========================================+
| rounding           | `round(x, digits = n)`                  | `round(3.5, digits = 0)`                  |
|                    |                                         |                                           |
|                    |                                         | `round(4.5, digits = 0)`                  |
+--------------------+-----------------------------------------+-------------------------------------------+
| rounding           | `janitor::round_half_up(x, digits = n)` | `janitor::round_half_up(3.5, digits = 0)` |
+--------------------+-----------------------------------------+-------------------------------------------+
| ceiling (round up) | `ceiling(x)`                            | `ceiling(3.1234)`                         |
+--------------------+-----------------------------------------+-------------------------------------------+
| floor (round down) | `floor(x)`                              | `floor(3.1234)`                           |
+--------------------+-----------------------------------------+-------------------------------------------+
| absolute value     | `abs(x)`                                | `abs(3.1234)`                             |
+--------------------+-----------------------------------------+-------------------------------------------+
| square root        | `sqrt(x)`                               | `sqrt(3.1234)`                            |
+--------------------+-----------------------------------------+-------------------------------------------+
| exponent           | `exponent(x)`                           | `exponent(3.1234)`                        |
+--------------------+-----------------------------------------+-------------------------------------------+
| natural logarithm  | `log(x)`                                | `log(3.1234)`                             |
+--------------------+-----------------------------------------+-------------------------------------------+
| log base 10        | `log10(x)`                              | `log10(3.1234)`                           |
+--------------------+-----------------------------------------+-------------------------------------------+
| log base 2         | `log2(x)`                               | `log2(3.1234)`                            |
+--------------------+-----------------------------------------+-------------------------------------------+

## Useful statistical functions

- If `x` is a vector of numeric values, then we can perform several statistical functions on it

| Objective               | Function             |
|-------------------------|----------------------|
| mean (average)          | `mean(x, na.rm=T)`   |
| median                  | `median(x, na.rm=T)` |
| standard deviation      | `sd(x, na.rm=T)`     |
| quantiles\*             | `quantile(x, probs)` |
| sum                     | `sum(x, na.rm=T)`    |
| minimum value           | `min(x, na.rm=T)`    |
| maximum value           | `max(x, na.rm=T)`    |
| range of numeric values | `range(x, na.rm=T)`  |
| summary\*\*             | `summary(x)`         |

***CAUTION:*** The functions above will by default include missing values in calculations. Missing values will result in an output of `NA`, unless the argument `na.rm = TRUE` is specified. This can be written shorthand as `na.rm = T`.

## Other useful functions

+----------------------+---------------------+---------------------------------------------+
| Objective            | Function            | Example                                     |
+======================+=====================+=============================================+
| create a sequence    | `seq(from, to, by)` | `seq(1, 10, 2)`                             |
+----------------------+---------------------+---------------------------------------------+
| repeat x, n times    | `rep(x, ntimes)`    | `rep(1:3, 2)` or `rep(c("a", "b", "c"), 3)` |
+----------------------+---------------------+---------------------------------------------+
| take a random sample | `sample(x, size)`   | `sample(1:15, size = 5, replace = TRUE)`    |
+----------------------+---------------------+---------------------------------------------+

::: blue
::: blue-cont
Extra note: we will "set a seed" for `sample()` because the process is random, and the output will change everytime. `set.seed()` will anchor us in a specific instance of the random number generator.
:::
:::

## Wrap-up

- We learned some boolean operators to expand our true/false logical work
- We learned *many* new functions
  - We'll get to practice them in our practice assignment!
  
 

- Main goal was to gain recognition of functions
- Hopefully, we can use what we learned in the lesson on getting help in R to investigate the functions more!

## Resources

- [This lesson's R script with the examples](./Key_operators_examples.R)
