---
title: "Homework 2 Answers"
subtitle: "BSTA 513/613"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html: 
    link-external-newwindow: true
    toc: true
    embed-resources: true
    self-contained: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

```{r setup, include=FALSE}

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
library(readxl)
library(haven)
library(here)
library(gtsummary) ## Added this package!!
library(kableExtra)
```

## Questions

### Question 1

```{r}
#| echo: false

icu<-read.csv("../data/icu.csv")
icu$STA <- ifelse(icu$STA=="Died",1,0)
icu$STA_f<-factor(icu$STA)
```


#### Part a

Write down the equation for the logistic regression model of STA on AGE. What characteristic of the outcome variable, STA, leads us to consider the logistic regression model as opposed to the usual linear regression model to describe the relationship between STA and AGE?

**Answer:**
<div class = "blue">
Not given
</div>

#### Part b

Write down an expression for the log-likelihood for the logistic regression model in Part a. This will me a mathematical expression. Please do not use generic expressions like $\pi(X)$, instead replace $X$ with the specific variables in this question. 

**Answer:**
<div class = "blue">

$$L(\beta_0,\beta_1) =  \ln(l(\beta_0,\beta_1))= \ldots$$

</div>

#### Part c

Using the `glm()` function, obtain the maximum likelihood estimates of the coefficient parameters of the logistic regression model in Part a. Using these estimates, write down fitted logistic regression model.

**Answer:**
<div class = "blue">
Coefficient estimates:
```{r}
#| echo: false

mylogit = glm(STA ~ AGE, data = icu, family = binomial)
tidy_age = tidy(mylogit, conf.int=T) 
tidy_age %>% gt() %>% tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```
</div>

#### Part d
Use the Wald test to test whether or not the intercept ($\beta_0$) of the logistic regression model is significantly different from 0. Make sure to include your: hypothesis test, test statistic, code/work leading to the computed test statistic, distribution of the test statistic, interpretation of the intercept, and conclusion. Please refer to the Additional Tips to guide you on what a complete/correct answer contains. 

**Answer:**
<div class = "blue">
Estimated odds and odds ratio:
```{r}
#| echo: false

tidy_age = tidy(mylogit, conf.int=T, exponentiate=T) 
tidy_age %>% gt() %>% tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```
</div>


#### Part e
Use the Likelihood Ratio test to test whether or not the coefficient for age ($\beta_1$) of the logistic regression model is significantly different from 0. Make sure to include your: hypothesis test, test statistic, code/work leading to the computed test statistic, distribution of the test statistic, and conclusion. Please refer to the Additional Tips to guide you on what a complete/correct answer contains. **You do not need to include an interpretation of the coefficient since we have not covered this.**

**Answer:**
<div class = "blue">
```{r}
#| echo: false
#| message: false
#| warning: false

library(lmtest)
age_glm = glm(STA ~ AGE, data = icu, family = binomial)
int_glm = glm(STA ~ 1, data = icu, family = binomial)

lrtest(age_glm, int_glm)
```
</div>

#### Part f

Write a sentence interpreting the estimated odds ratio for the coefficient in Part e. Please include the 95% confidence interval.

::: blue
Estimated odds and odds ratio for the fitted model:
```{r}
#| echo: false

tidy_age = tidy(age_glm, conf.int=T, exponentiate = T) 
tidy_age %>% gt() %>% tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```
:::

#### Part g {#sec-L06-OR-interp}

Write a sentence interpreting the estimated odds ratio for a 10-year increase in age. Please include the 95% confidence interval. (Hint: We're scaling up the coefficient 10 times)

**Answer:**

::: blue

Estimated odds ratio: 

```{r}
#| echo: false

icu_new = icu %>% mutate(AGE_10 = AGE/10)
age10_glm = glm(STA ~ AGE_10, data = icu_new, family = binomial)
tidy_age10 = tidy(age10_glm, conf.int=T, exponentiate = T) 
tidy_age10$estimate[2]
```

:::

#### Part h {#sec-L07-pred-prob}

Compute the predicted probability of hospital discharge for a subject who is 63 years old. Compute the 95% confidence interval for the predicted probability and interpret the predicted probability.

**Answer:**
<div class = "blue">

With Normal approximation:
```{r}
#| echo: false

df = data.frame(AGE=63)
prop = predict(age_glm, df, se.fit = T, 
               type="response") # get estimated probability directly
```

```{r}
#| echo: false

lower <- prop$fit - 1.96*prop$se.fit
upper <- prop$fit + 1.96*prop$se.fit


t2 = cbind("Probability",round(prop$fit,digits = 4),round(lower,digits = 4),round(upper,digits = 4))

pred = c(Prediction = prop$fit, lower, upper)

library(boot)
pred %>% round(., digits=3)
```

</div>

### Question 2

We will continue to work with the ICU data in Question 1. Please refer back to the information above. In this question, we will use the ICU data to fit a multivariable logistic regression model. 


#### Part a

From the above list (AGE, CAN, CPR, INF, and LOC) of independent variables, identify if each is a continuous, binary, or multi-level (\>2) categorical variable.

**Answer:**
<div class = "blue">
Continuous: ??

Binary: CAN, CPR, ???

Multi-level (>2) categorical: ??
</div>

#### Part b

For the binary and multi-level categorical variables, please identify a reference group for each. Include justification for the reference group.

**Answer:**
<div class = "blue">
Not given - just need some thought
</div>



#### Part c

For the categorical variables (binary and multi-group), please mutate the variables within the ICU dataset to set your chosen reference groups.

**Answer:**
<div class = "blue">
Not given

```{r}
#| echo: false
icu = icu %>% mutate(CAN = as.factor(CAN) %>% relevel(ref = "No"), 
                     CPR = as.factor(CPR) %>% relevel(ref = "No"), 
                     INF = as.factor(INF) %>% relevel(ref = "No"), 
                     LOC = as.factor(LOC) %>% 
                       relevel(ref = "No Coma or Deep Stupor"))
```
</div>

#### Part d

Write down the equation for the logistic regression model of STA on CPR.

**Answer:**
<div class = "blue">
$$\text{logit} \left(\pi(CPR)\right)= \ldots$$
</div>

#### Part e

Using the `glm()` function, obtain the maximum likelihood estimates of the coefficient parameters of the logistic regression model in Part f. Using these estimates, write down fitted logistic regression model.

**Answer:**
<div class = "blue">

```{r}
#| echo: false
cpr_glm = glm(STA ~ CPR, data = icu, family = binomial)
cpr_coef = tidy(cpr_glm, conf.int = T)
cpr_coef %>% gt() %>% 
  tab_options(table.font.size = 16) %>%
  fmt_number(decimals = 3)
```

$$\begin{aligned}
\text{logit} \left(\widehat{\pi}(CPR)\right) &= `r round(cpr_coef$estimate[1], 3)` + \ldots
\end{aligned}$$

</div>

#### Part f

Write a sentence interpreting the odds ratio for the coefficients in Part g's model. Please include the 95% confidence interval.

**Answer:**
<div class = "blue">

```{r}
#| echo: false

cpr_OR = tidy(cpr_glm, conf.int = T, exponentiate = T)
cpr_OR %>% gt() %>% 
  tab_options(table.font.size = 16) %>%
  fmt_number(decimals = 3)
```

OR PRESENT THIS WAY:
```{r}
#| echo: false

library(gtsummary)
tbl_regression(cpr_glm, exponentiate = TRUE) %>% 
  as_gt() %>% # allows us to use tab_options()
  tab_options(table.font.size = 16)
```

</div>

#### Part g

Write down the equation for the logistic regression model of STA on LOC.

**Answer:**
<div class = "blue">
$$\text{logit} \left(\pi(LOC)\right)= \ldots$$
</div>

#### Part h

Using the `glm()` function, obtain the maximum likelihood estimates of the coefficient parameters of the logistic regression model in Part h. Present the coefficient estimates. No need to write out the fitted regression equation. 

Please take note of the warnings that you receive from fitting the `glm()` model and any large coefficient estimate with large confidence intervals. In this case, we have a category within LOC that has very few observations.

Check the number of observations that have a deep stupor and death at discharge and the number of observations that have a deep stupor and live at discharge. You can do this using the `table()` function to create a contingency table. 

**Answer:**
<div class = "blue">
```{r}
#| echo: false
#| warning: false
#| message: false

loc_glm = glm(STA ~ LOC, data = icu, family = binomial)
loc_coef = tidy(loc_glm, conf.int = T)
loc_coef %>% gt() %>% 
  tab_options(table.font.size = 16) %>%
  fmt_number(decimals = 3)
```
</div>

#### Part i

Write a sentence interpreting the odds ratio of death for the indicator of coma. Please include the 95% confidence interval.

**Answer:**
<div class = "blue">
```{r}
#| echo: false
#| warning: false
#| message: false

loc_OR = tidy(loc_glm, conf.int = T, exponentiate = T)
loc_OR %>% gt() %>% 
  tab_options(table.font.size = 16) %>%
  fmt_number(decimals = 3)
```
</div>


