---
title: "Practice 4 Answers"
subtitle: "PUBH 523/623"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
    highlight-style: ayu
    code-block-bg: "#f8f9fa"   
    code-block-border-left: "#5BAFF8" 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

## Questions

### Question 1: Import and Export

In Lesson 18, we learned how to import and export data using file-specific functions and the `rio` package.

#### Part A: Import the HRS dataset

::: {.callout-warning appearance="minimal"}
##### Task

Load the `tidyverse`, `rio`, and `here` packages. Then use `import()` together with `here()` to import `hrs_data.rds` from your `data` folder. Assign it to an object called `hrs_data`. Use `glimpse()` (or `tibble()`) to take a look at its structure.
:::

::: blue
**Answer:**

```{r}
pacman::p_load(
  tidyverse, 
  rio, 
  here
)

hrs_data <- import(here("data", "hrs_data.rds")) #<1>
```

1. This file path is specific to my folder set-up. 
:::

#### Part B: Export and re-import

::: {.callout-warning appearance="minimal"}
##### Task

1. Use `export()` from `rio` to save `hrs_data` as a `.csv` file named `hrs_data_export.csv` in the same folder as the `hrs_data.rds` file. 
2. Use `import()` from `rio` to read `hrs_data_export.csv` back into R as a new object called `hrs_data_csv`.
3. Use `class()` to check the class of the `sex` column in **both** `hrs_data` and `hrs_data_csv`. In 1-2 sentences, describe what you notice — is the class the same or different, and why might that be?
:::

::: blue
**Answer:**

Not given
:::

#### Part C: When to use which file type

::: {.callout-warning appearance="minimal"}
##### Task

In 1-2 sentences, explain when you would prefer to save a dataset as `.rds` versus `.csv`, based on what we discussed in Lesson 18.
:::

::: blue
**Answer:**

Not given
:::

### Question 2: Piping

In Lesson 16, we learned that pipes (`|>`) let us write code that reads left to right, passing the output of one step into the next.

#### Part A: Nested to piped

::: {.callout-warning appearance="minimal"}
##### Task

Here is a chunk of **nested** code:

```{r}
#| eval: false
paste0("Mean CESD score: ", round(mean(hrs_data$cesd, na.rm = TRUE), digits = 1))
```

Rewrite this same calculation using the pipe (`|>`) instead of nesting. You will need the placeholder `_` for at least one step, since `paste0()` does not take the piped input as its first argument. Assign the result to an object called `cesd_summary` and print it.
:::

::: blue
**Answer:**

```{r}
cesd_summary <- hrs_data$cesd |> 
  mean(na.rm = TRUE) |> 
  round(digits = 1) |> 
  paste0("Mean CESD score: ", x = _)

cesd_summary
```
:::

#### Part B: Write your own pipe

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data$height`, write a pipe that (1) calculates the mean while handling any missing values, and (2) rounds that mean to 1 decimal place. Assign the result to an object called `height_summary` and print it.
:::

::: blue
**Answer:**

Not given
:::

### Question 3: Key operators and functions

In Lesson 17, we expanded on boolean operators and were introduced to several new functions.

#### Part A: Boolean operators and `%in%`

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data$sex`, answer the following with R code (don't just eyeball the data frame):

1. Is `"Female"` one of the values in `hrs_data$sex`? Use `%in%`.
2. Write one logical statement that uses `&` to check whether the **first** respondent (`hrs_data$age_yr[1]`) is over 65 **and** the **first** respondent's `sex` is `"Female"`.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: Statistical functions

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data$cesd`, calculate each of the following and print them: the standard deviation (`sd()`), the range (`range()`), and the mean rounded to the nearest whole number using `round()`. (Handle any missing values you find along the way.)
:::

::: blue
**Answer:**

Standard deviation result:
```{r}
#| echo: false
sd(hrs_data$cesd, na.rm = TRUE)
```
:::

#### Part C: `seq()`, `rep()`, and `sample()`

::: {.callout-warning appearance="minimal"}
##### Task

1. Use `set.seed(523)` and then `sample()` to randomly draw 5 ids (without replacement) from `hrs_data$id`. Assign this to `sampled_ids` and print it.
2. In 1 sentence, explain why we use `set.seed()` before `sample()`.
:::

::: blue
**Answer:**

Sampled ID's: 
```{r}
#| echo: false

set.seed(523)
sample(hrs_data$id, size = 5, replace = FALSE)
```
:::

### Question 4: Tidy data

In Lesson 19, we learned that tidy data follow three rules: each variable has its own column, each observation has its own row, and each value has its own cell.

#### Part A: Is it tidy?

::: {.callout-warning appearance="minimal"}
##### Task

Below is a small dataset showing the same information two different ways. For **each** table, state whether it is tidy or not, and explain why in 1-2 sentences (referencing the three rules above).

**Table 1**

| id | age_yr | cesd |
|----|--------|------|
| R1 | 55     | 2    |
| R2 | 62     | 4    |

**Table 2**

| id | variable | value |
|----|----------|-------|
| R1 | age_yr   | 55    |
| R1 | cesd     | 2     |
| R2 | age_yr   | 62    |
| R2 | cesd     | 4     |
:::

::: blue
**Answer:**

Table 1 is tidy. Table is not tidy. 
:::

#### Part B: Is `hrs_data` tidy?

::: {.callout-warning appearance="minimal"}
##### Task

Look at `hrs_data`. In 1-2 sentences, explain whether it is tidy, and identify what one row and one column represent.
:::

::: blue
**Answer:**

Not given
:::
