---
title: "Practice 3 Answers"
subtitle: "PUBH 523/623"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

## Questions

### Question 1: Getting to Know the HRS Dataset

In Lesson 11, we were introduced to the *synthetic* HRS (Health and Retirement Study) dataset we will use for the rest of the course.

#### Part A: Explore the codebook

::: {.callout-warning appearance="minimal"}
##### Task

Open the [HRS codebook](https://nwakim.github.io/PUBH_523_26Su/resources/HRS_codebook.html) and answer the following:

1.  What does the variable `srh` measure, and what are its possible levels?
2.  Name **two** variables in the dataset that are stored as `factor`, and **two** that are stored as `numeric`.
3.  In 1–2 sentences, explain why this is a *synthetic* dataset rather than the real HRS data, and what that means for any patterns you might see in it.
:::

**Answer:**

::: blue
No answer given, for completion only.
:::

#### Part B: A mini HRS sample

```{r}
#| message: false
pacman::p_load(tidyverse)

mini_hrs <- data.frame(
  id     = c("R1", "R2", "R3", "R4", "R5"),
  age_yr = c(55, 62, 71, 48, 67),
  sex    = c("Female", "Male", "Male", "Female", "Male"),
  cesd   = c(2, 4, 6, 0, NA)
)

mini_hrs
```

::: {.callout-warning appearance="minimal"}
##### Task

Compare `mini_hrs` to the [HRS codebook](https://nwakim.github.io/PUBH_523_26Su/resources/HRS_codebook.html). In 2–3 sentences, describe what one row of `mini_hrs` represents, and name one way this mini sample is similar to (or different from) the full synthetic HRS dataset described in Lesson 11.
:::

**Answer:**

::: blue
Not given
:::

### Question 2: Functions

In Lesson 12, we learned that functions take the general form `function_name(argument1 = value1, argument2 = value2, ...)`.

#### Part A: Call a function with arguments

::: {.callout-warning appearance="minimal"}
##### Task

Use the `seq()` function to generate a sequence of even numbers from 2 to 20, naming your arguments (`from`, `to`, `by`) explicitly. Assign the output to an object called `evens`, then print `evens`.
:::

**Answer:**

::: blue
The resulting sequence should look like:

```{r}
#| echo: false
evens <- seq(from = 2, to = 20, by = 2)
evens
```
:::

#### Part B: Use a function on your data

::: {.callout-warning appearance="minimal"}
##### Task

Using your `mini_hrs` data frame from Question 1, calculate the mean of the `cesd` column using the `mean()` function. Access the column with the `$` operator (Lesson 13).
:::

**Answer:**

::: blue
No answer given, for completion only.
:::

#### Part C: Look up a function's help file

::: {.callout-warning appearance="minimal"}
##### Task

Pick any **one** function you used in Question 1 or 2 (for example, `data.frame()`, `seq()`, or `mean()`). Open its help file using `?`. In 2–3 sentences, describe what you found: what are the function's main arguments, and is there anything in the help file that surprised you or that you didn't know before?
:::

**Answer:**

::: blue
Not given
:::

### Question 3: Objects

In Lesson 13, we learned about different types of R objects, both single pieces of data and collections of data.

#### Part A: Object types

::: {.callout-warning appearance="minimal"}
##### Task

Fill in the table below with an example value for each object class, and one sentence describing when you might use that type in the HRS dataset.
:::

**Answer:**

::: blue
| Class | Example value | When might you see this in the real HRS dataset (see codebook)? |
|---------------|---------------|------------------------------------------|
| Double (`dbl`) | `71` or `3.5` | Continuous or count measures, like `age_yr` or a summed depression score |
| Character (`chr`) |  |  |
| Factor (`fct`) |  |  |
| Logical (`lgl`) |  |  |

Not given
:::

#### Part B: Check the class of variables

::: {.callout-warning appearance="minimal"}
##### Task

Using your `mini_hrs` data frame, use the `class()` function to check the object type of each of its four columns. Access each column with the `$` operator.
:::

**Answer:**

::: blue
```{r}
class(mini_hrs$id)
class(mini_hrs$age_yr)
```
:::

### Question 4: Common Errors

Even experienced R users run into errors constantly! Lesson 14 covered some of the most common ones.

#### Part A: Spot and fix the bugs

::: {.callout-warning appearance="minimal"}
##### Task

Each code chunk below contains **one error**. For each chunk, (1) identify what is causing the error, and (2) fix the code so it runs correctly.
:::

```{r}
#| eval: false
# Chunk 1
mean(mini_hrs$Cesd)
```

::: blue
**Answer:** `cesd` is case-sensitive, so it should be `mini_hrs$cesd` instead of `mini_hrs$Cesd`
:::

```{r}
#| eval: false
# Chunk 2
avg_age <- mean(mini_hrs$age_yr
```

::: blue
**Answer:** Not given
:::

```{r}
#| eval: false
# Chunk 3
ages <- c(55, 62, 71, 48, 67)
avg <- mean("ages")
```

::: blue
**Answer:** Not given
:::

#### Part B: Read an error message

::: {.callout-warning appearance="minimal"}
##### Task

We ran the following code chunk and got the printed error. In 1–2 sentences, explain what the error message is telling you.
:::

```{r}
#| error: true
mini_hrs$age_yr + mini_hrs$id
```

**Answer:**

::: blue
Not given
:::

### Question 5: Getting Help

Knowing how to get *unstuck* is one of the most important skills in R. Lesson 15 covered strategies for getting help.

#### Part A: Search for your error

::: {.callout-warning appearance="minimal"}
##### Task

Take the error message from Question 4 Part B. Search for it online (Google, Stack Overflow, or another resource). Link to a resource you found helpful, and in 1–2 sentences summarize what it suggested.
:::

**Answer:**

::: blue
Not given
:::

#### Part B: Ask a good question

::: {.callout-warning appearance="minimal"}
##### Task

Imagine you still can't solve the error from Question 4 Part B, and you need to email your instructor for help. Write the email you would send. Your email should include the pieces of information a good help-seeking question needs (think about what you'd want to know if you were the one answering).
:::

**Answer:**

::: blue
Not given
:::
