---
title: "Data transformation: Rows, *Advanced*"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: Rows, *Advanced*"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readxl)
```

```{r}
#| echo: false
hrs_data <- read_excel(here("./data/hrs_data.xlsx"))
hrs_data <- clean_names(hrs_data)
```

# 

## Filtering requires a little logic

-   We can chain multiple criteria using the [`&` (AND)]{style="color:darkorange"} or [`|` (OR) operators]{style="color:darkorange"}.
    -   But we need to review a little logic before we do this.



-   If we wanted to return participants who identified as
    -   `Female` **and** are in 10th grade,
    -   we would use an `&` to chain these criteria together:

```         
sex == "Female" & grade == "10th"
```



-   Note that *SPELLING AND CASE MATTER*!
    -   You must write the name of the column exactly and
    -   also the value (i.e. "Female" not "female") exactly.

## Filtering with `&` (AND) criteria

```{r}
hrs_data |>
  filter(diab == "Yes" & 
           degree == "Master's degree")

```

-   Note that we could also use the comma in this case:

```{r}
hrs_data |>
  filter(diab == "Yes",    # <1> Comma is equivalent to "&" within filter()
         degree == "Master's degree")

```

## Filtering with `|` (OR) criteria

If we wanted patients who were

`Female` [**or**]{style="color:darkorange"} in 10th grade

we would use a `|` to chain these criteria together.

```{r}
hrs_data |>
  filter(diab == "Yes" |
           degree == "Master's degree")
```

::: {.callout-warning icon="false"}
## Think about it:

Which of the AND vs. OR code blocks will return a larger number of participants?
:::

## More filter examples

-   See [this slide](https://jminnier-berd-r-courses.netlify.app/02-data-wrangling-tidyverse/02_data_wrangling_slides_part1.html#27){target="_blank"} from a previous BERD workshop.

-   Also read the [filter section](https://r4ds.hadley.nz/data-transform.html#rows){target="_blank"} in the "Data Transformation" chapter from *R for Data Science (2e)*.

<br>

### More about comparison and logical operators

Click [here](https://www.datamentor.io/r-programming/operator/){target="_blank"} for a useful reference for all the different operators (both logical and comparison) that you can use.

## 

Using %in%

# `arrange()`

## `arrange()` after `count()`-ing categorical data

-   Another example of when `arrange()` is useful

::: columns
::: {.column width="48%"}
-   We can also use `count()` to count all of the unique values for a categorical variable.
-   This will also show `NA` category, which is the special missing variable type:

```{r}
hrs_data |>
  count(age)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
-   Sometimes it helps to arrange by that category if you have many categories to view.
-   This can also alert us if there are categorical coding mistakes (such as misspellings) in our data.

```{r}
hrs_data |>
  count(age) |>
  arrange(n)
```
:::
:::