---
title: "Intro to data transformations and summarizations"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Intro to data transformations and summarizations"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl
)
```

```{r}
#| echo: false

hrs_data <- read_excel(here("./data/hrs_data.xlsx"))
hrs_data <- clean_names(hrs_data)
```

## `dplyr` basics


- `dplyr` is a package in the tidyverse that provides a **set of functions for data manipulation and transformation**

 

- **Each function** in `dplyr` is like a **verb that describes the action** being performed on the data

 

- Most functions within `dplyr` have similar input and output:
  - *Input:* a data frame (usually the first argument in the function)
  - *Output:* a data frame
  - *Additional arguments:* typically describe what columns (variables) to operate on

## Example of a `dplyr` function: `select()`

::: columns
::: {.column width="30%"}

::: pink2
::: pink2-cont
I can use the `select()` function to select specific columns (`age` and `ed`) from my data frame (`hrs_data`)
:::
:::
:::
::: {.column width="5%"}
:::
::: {.column width="65%"}
```{r}
hrs_data |> # <1> 
  select( # <2>
    age, # <3>
    ed  # <3>
    ) # <2>
```

1. The first argument is the data frame we want to operate on (in this case, `hrs_data`)
2. `select` is the function we are using on the data frame
3. `age` and `ed` are the columns (in the data frame) that we perform the `select()` function on

:::
:::

## Types of `dplyr` functions

`dplyr` functions can be categorized into four groups based on what they operate on:

1. [**rows**]{style="color: #5BAFF8;"}: functions that operate on rows of a data frame 
    - Examples: `filter()`, `slice()`
    - Under the Data Transformation lesson series
2. [**columns**]{style="color: #E75B5C;"}: functions that operate on columns of a data frame
    - Examples: `select()`, `mutate()`
    - Under the Data Transformation lesson series
3. [**groups**]{style="color: #EF85B3;"}: functions that operate on groups of rows in a data frame 
    - Examples: `group_by()`, `summarize()`
    - Under the Data Summarization lesson series
4. [**joins**]{style="color: #459B99;"}: functions that combine two data frames based on a common key 
    - Examples: `left_join()`, `inner_join()`
    - Not covered in this class

## Resources

- [`dplyr` cheat sheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)

