---
title: "Import and Export"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Import and Export"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  janitor,
  here, 
  rio, 
  readxl
)
```

```{r}
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_data <- clean_names(hrs_data_rds)
```

## Importing data

- We often need to work with a dataset in R
  - These data need to be brought into our environment so we can work with them!

 

- We can import data from many file types, including `.csv`, `.txt`, `.xlsx`, `.rds`, `.sas7bdat`, and `.dta`

 

- Once imported, R typically stores data as **data frames** (or **tibbles**) if using the `{tidyverse}` package

  - Usually use the term **dataset** to generally refer to either a data frame or tibble

## Functions to import different file types

Here are various functions needed to import data from common file types:

| File Type | Extension(s) | Function | Package | Example |
|-------------|-------------|-------------|-------------|----------------------|
| CSV | `.csv` | `read.csv()`, `read_csv()` | base R, readr | `read.csv("file.csv")` , `read_csv("file.csv")` |
| Excel | `.xlsx`, `.xls` | `read_excel()` | readxl | `read_excel("file.xlsx", sheet = 1)` |
| RDS | `.rds` | `read_rds()` | readr | `read_rds("file.rds")` |
| SAS | `.sas7bdat` | `read_sas()` | haven | `read_sas("file.sas7bdat")` |
| Stata | `.dta` | `read_dta()` | haven | `read_dta("file.dta")` |
| SPSS | `.sav` | `read_sav()` | haven | `read_sav("file.sav")` |

## Let's see an example with our HRS data

- Notice I am using the `here` package to make a relative file path for my data

- If my HRS dataset is a `.rds` file type

  - Note: I prefer `.rds` files when available because it can store factors and other R-specific information

  ```{r}
  hrs_data_rds <- read_rds(here("./data/hrs_data.rds"))
  ```

 

- If my HRS dataset is a `.xlsx` file type

```{r}
hrs_data_xlsx <- read_excel(here("data", "hrs_data.xlsx"))
```

## Glimpse at the data frame

::: panel-tabset
### Data frame imported from `.rds`

```{r}
tibble(hrs_data_rds) |> print(n = 7)
```

### Data frame imported from `.xlsx`

```{r}
tibble(hrs_data_xlsx) |> print(n = 7)
```
:::

## `rio` package

- The `rio` package provides a single, universal import/export function that works across many file types
- Instead of remembering which package and function to use, `import()` detects the file type automatically from the extension

```{r}
#| eval: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))  #<1>
hrs_data_xlsx <- import(here("data", "hrs_data.xlsx")) #<2>
```

1.  This will automatically use `read_rds` to import the dataset
2.  This will automatically use `read_excel` to import the dataset

- Great for quickly reading in data without worrying about the right package

 

::: pink
::: pink-cont
Note: this is a new package! You will need to install it!
:::
:::

## Optional: cleaning names (1/2)

- Variable names in real datasets are often messy: spaces, capital letters, special characters
- The `clean_names()` function from `janitor` standardizes all column names

```{r}
names(hrs_data_rds) |> head(10) #<1>
```

1.  Original names from our imported HRS dataset

```{r}
hrs_data <- clean_names(hrs_data_rds) #<2>
names(hrs_data) |> head(10) #<3>
```

2.  We can use `clean_names()` on the dataset directly
3.  New, clean names for the HRS dataset

 

- Spaces are converted to underscores: "Age (years)" to "age_yrs"
- Upper case letters are converted to lower case: "BIRTHDATE" to "birthdate"

## Optional: cleaning names (2/2)

- Consistent, clean names make coding easier and reduce errors

 

- Can be piped directly after import:

```{r}
hrs_data <- import(here("data", "hrs_data.rds")) |> clean_names()
```

## Exporting data

- After cleaning or modifying data, you may want to save it to a file
- Use `export()` from `rio` to write data to any common format:

```{r}
#| eval: false
export(hrs_data, here("data", "hrs_data_clean.rds"))   #<1>
export(hrs_data, here("data", "hrs_data_clean.csv"))   #<2>
export(hrs_data, here("data", "hrs_data_clean.xlsx"))  #<3>
```

1. Save a `.rds` file in the data folder
2. Save a `.csv` file in the data folder
3. Save a `.xlsx` file in the data folder

- Prefer `.rds` for saving R objects between sessions: it preserves variable types exactly
- Use `.csv` or `.xlsx` when sharing data with others outside of R

## There are also the equivalent export functions

Here are various functions needed to **export** data from common file types:

| File Type | Extension(s)    | Function                          | Package          | Example                                                    |
|----------------|-----------------|-----------------------------------|------------------|--------------------------------------------------------|
| CSV       | `.csv`          | `write.csv()`, `write_csv()`      | base R, readr    | `write.csv(df, "file.csv")`, `write_csv(df, "file.csv")`  |
| Excel     | `.xlsx`         | `write_xlsx()`                    | writexl         | `write.xlsx(df, "file.xlsx")`                              |
| RDS       | `.rds`          | `write_rds()`                     | readr            | `write_rds(df, "file.rds")`                                |
| SAS       | `.sas7bdat`     | `write_sas()`                     | haven            | `write_sas(df, "file.sas7bdat")`                           |
| Stata     | `.dta`          | `write_dta()`                     | haven            | `write_dta(df, "file.dta")`                                |
| SPSS      | `.sav`          | `write_sav()`                     | haven            | `write_sav(df, "file.sav")`                                |

## Wrap-up

- Today we covered how to **import** and **export** data in R from a variety of file types

 

- Key takeaways:
  - Use file-specific functions (`read_csv()`, `read_excel()`, etc.) or `import()` from `rio` for convenience
  - Always **check variable types** after importing with `glimpse()` or `tibble()`
  - Use `clean_names()` from `janitor` to standardize column names
  - Use `export()` or file-specific functions to **save** cleaned data

## Resources

- `readr` cheatsheet
  - [Website form](https://rstudio.github.io/cheatsheets/html/data-import.html)
  - [PDF form](https://rstudio.github.io/cheatsheets/data-import.pdf)
