---
title: "Lesson 4: Calling Functions"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 03 Slides
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: below
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

# Learning Objectives

## What are functions?

- R has many built-in functions that allow you to perform a variety of tasks

- Functions usually take the form:

```{r}
#| eval: false

function_name(                           # <1>
  argument1 = input_value1,              # <2>
  argument2 = input_value2,              # <2>
  ...                                    # <2>
  )                                      # <1>
```

1. `function_name` is the name of the function we are using. `function_output` 
2. Within the parentheses of the function, there are set arguments that we assign values to. 

## Example: calling the `seq()` function

- `seq()` is a function that generates a sequence of numbers. 
- It has three arguments: `from`, `to`, and `by`
- Output will be the sequence numbers 

```{r}
seq(
  from = 1, 
  to = 10, 
  by = 2
)
```

## We don't need to specify the argument names

```{r}
seq(1, 10, 2)
```

## Output of a function

- We can also assign the output of a function to an object using the assignment operator `<-`:

```{r}
#| eval: false

function_output <- function_name(        # <1>
  argument1 = input_value1,              # <2>
  argument2 = input_value2,              # <2>
  ...                                    # <2>
  )                                      # <1>
```

- The output will not print, but it will be stored in the object `function_output`

## Now with the `seq()` function

- We will assign the output of the `seq()` function to an object called `odds`:

```{r}
odds <- seq(
  from = 1, 
  to = 10, 
  by = 2
)
```

- So `odds` will now contain the sequence (1,3,5,7,9)

