---
title: "Common errors for data transformations"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Common errors for data transformations"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  # janitor,
  here
)
```

## Types of common errors

1.  [Typo errors]{style="color:#BF396F;"}
2.  [Package errors]{style="color:#3070BF;"}
3.  [Object errors]{style="color:#459B99;"}
4.  [Function syntax errors]{style="color:#C83532;"}
5.  [Logic errors]{style="color:#BF396F;"}
6.  [Factor errors]{style="color:#3070BF;"}
7.  [Plotting errors]{style="color:#459B99;"}
8.  [R Markdown / Quarto errors]{style="color:#C83532;"}

:::: pink
::: pink-cont
We will go through each type with real error messages and how to fix them. Keep this slide as a reference!
:::
::::


## [1. Typo errors]{style="color:#BF396F;"}

::::::: columns
::::: {.column width="50%"}
**Missing comma**

```         
Error: unexpected symbol in:
"  geom_histogram(stat = "identity")+
  tidyquant::geom_ma(n=7, size = 2, color = "red" lty"
```

:::: pink
::: pink-cont
If you see **"unexpected symbol"**, the first thing to check is a **missing comma**.
:::
::::

R is case-sensitive — `Mean()` and `mean()` are different things!

```{r}
#| error: true
Mean(c(1, 2, 3))   #<1>
```

1.  Capital M — R can't find this function
:::::

::: {.column width="50%"}
**Mismatched parentheses**

```{r}
#| error: true
mean(c(1, 2, 3)   #<1>
```

1.  Missing closing `)` — the console shows `+` waiting for more input

**Quick fixes:**

- Use **Tab** to autocomplete function names and avoid typos
- Check that every `(` has a closing `)`
- Use RStudio's bracket highlighting — click next to a `(` and its match will highlight
- Press **Escape** if the console gets stuck on `+`
:::
:::::::

------------------------------------------------------------------------

## [2. Package errors (1/2)]{style="color:#3070BF;"}

**Function not found**

```         
could not find function "x"...
```

This usually means you typed the function name incorrectly, or **forgot to install or load the package**.

:::::::: columns
::: {.column width="50%"}
```{r}
#| error: true
clean_names(iris)   #<1>
```

1.  `clean_names()` is from `janitor` — if it's not loaded, R can't find it

✅ Fix:

```{r}
library(janitor)       #<1>
clean_names(iris) |> names()
```

1.  Load the package first, then call the function
:::

:::::: {.column width="50%"}
::::: blue
::: blue-ttl
Install once, load every session
:::

::: blue-cont
- `install.packages("janitor")` — run **once**, downloads to your computer
- `library(janitor)` — run **every session**, makes it available in R

Think of it like: install = buy the book; library = open it.
:::
:::::
::::::
::::::::

------------------------------------------------------------------------

## [2. Package errors (2/2)]{style="color:#3070BF;"}

**Masked functions — when two packages have the same function name**

```         
Error in select(data, var) : unused argument (var)
```

:::::::: columns
::: {.column width="55%"}
This is a classic error! Both `dplyr` and `MASS` have a `select()` function. Whichever package was loaded **last** wins — and if `MASS` was loaded after `dplyr`, you'll get the wrong `select()`.

```{r}
#| eval: false
library(MASS)
library(dplyr)
iris |> select(Species)   #<1>
```

1.  Which `select()` is this? Depends on load order!

✅ Fix — be explicit with `package::function()`:

```{r}
iris |> dplyr::select(Species) |> head(3)   #<1>
```

1.  No ambiguity — always uses `dplyr`'s version
:::

::: {.column width="5%"}
:::

::::: {.column width="40%"}
**Other common masking conflicts:**

| Package A | Package B | Function      |
|-----------|-----------|---------------|
| `dplyr`   | `MASS`    | `select()`    |
| `dplyr`   | `plyr`    | `summarise()` |
| `dplyr`   | `stats`   | `filter()`    |

:::: blue
::: blue-cont
The [**conflicted** package](https://www.tidyverse.org/blog/2018/06/conflicted/) forces R to throw an error whenever a masked function is used — great for catching these silently wrong results.
:::
::::
:::::
::::::::

------------------------------------------------------------------------

## [3. Object errors]{style="color:#459B99;"}

::::: columns
::: {.column width="50%"}
**Object not found**

```         
object 'x' not found
```

This means R can't find a variable you referenced. Common causes:

```{r}
#| error: true
my_summary   #<1>
```

1.  This object was never created — code above may not have run

```{r}
#| error: true
install.packages(dplyr)   #<1>
```

1.  Missing quotes — R looks for an object named `dplyr`

✅ Fix: `install.packages("dplyr")`
:::

::: {.column width="50%"}
**File not found**

```         
No such file or directory
```

```{r}
#| error: true
#| eval: false
read_csv("my data.csv")   #<1>
```

1.  Typo, wrong path, or file doesn't exist yet

**Check:**

- Spelling of the file name
- Forward slashes `/` not backward `\`
- Correct file extension (`.csv`, `.xlsx`, `.rds`)
- That you are in the right **R Project**

**Subscript out of bounds**

```         
Error in 'x': subscript out of bounds
```

You tried to access an element that doesn't exist — e.g., asking for the 10th element of a 5-element vector.
:::
:::::

------------------------------------------------------------------------

## [4. Function syntax errors]{style="color:#C83532;"}

**Wrong or missing argument names**

:::::::: columns
::: {.column width="50%"}
A very common `mutate()` mistake — forgetting to repeat the column name inside `recode()`:

```{r}
#| error: true
iris |>
  mutate(Species = recode("setosa" ~ "S"))   #<1>
```

1.  `recode()` needs the column name as its first argument

✅ Fix:

```{r}
iris |>
  mutate(Species = recode(Species,
                          "setosa" = "S")) |>   #<1>
  count(Species)
```

1.  Pass `Species` as the first argument, then `OLD = NEW` pairs
:::

:::::: {.column width="50%"}
**Wrong argument name**

```{r}
#| error: true
mean(c(1, 2, NA), remove.na = TRUE)   #<1>
```

1.  `remove.na` is not a real argument — R ignores it and NA propagates

✅ Fix:

```{r}
mean(c(1, 2, NA), na.rm = TRUE)   #<1>
```

1.  The correct argument is `na.rm`

::::: orange
::: orange-ttl
How to find the right argument names
:::

::: orange-cont
Run `?function_name` and read the **Arguments** section — it lists every accepted argument and what it expects.
:::
:::::
::::::
::::::::

------------------------------------------------------------------------

## [5. Logic errors]{style="color:#BF396F;"}

**These don't always produce an error — the code runs, but gives you the wrong answer!**

::::::: columns
::: {.column width="50%"}
**`if` statement errors**

```         
Error in if
```

This usually means an `if` statement was applied to something that wasn't `TRUE` or `FALSE`:

```{r}
#| error: true
x <- c(1, 2, 3)
if (x > 1) print("big")   #<1>
```

1.  `x > 1` returns a *vector* — `if` expects a single TRUE/FALSE

✅ Fix — use `any()` or `all()`:

```{r}
if (any(x > 1)) print("at least one is big")   #<1>
```

1.  `any()` collapses the vector to a single TRUE/FALSE
:::

::::: {.column width="50%"}
**Silent logic errors (no error message!)**

```{r}
ages <- c(23, 31, 40, 45, 28)
mean(ages[ages = 30])   #<1>
```

1.  `=` should be `==` — this silently does something wrong!

✅ Fix:

```{r}
mean(ages[ages == 30])   #<1>
```

1.  `==` is "equal to"; `=` is assignment

:::: pink
::: pink-cont
Logic errors are the sneakiest bugs. Always **check intermediate results** with `print()`, `head()`, or `View()` to confirm your data looks right at each step.
:::
::::
:::::
:::::::

------------------------------------------------------------------------

## [6. Factor errors]{style="color:#3070BF;"}

**Factors look like character strings, but they have fixed levels — adding new values creates `NA`!**

::::::: columns
::::: {.column width="50%"}
**Adding a value not in the levels:**

```{r}
#| warning: true
age_cat <- factor(c("child", "adult", "elder"))
age_cat[1] <- "infant"   #<1>
age_cat
```

1.  "infant" is not a defined level — `NA` is introduced

:::: blue
::: blue-cont
This error message gives it away:

```         
invalid factor level, NA generated
```

If you see it, you tried to add a value that wasn't in the factor's pre-defined levels.
:::
::::
:::::

::: {.column width="50%"}
✅ Fix — convert to character first, then back to factor:

```{r}
age_cat <- factor(c("child", "adult", "elder"))
age_cat <- as.character(age_cat)   #<1>
age_cat[1] <- "infant"
age_cat <- factor(age_cat)         #<2>
levels(age_cat)
```

1.  Convert to character so any value is accepted
2.  Re-factor with all levels present

**Unexpected alphabetical ordering:**

```{r}
survey <- data.frame(
  rating = factor(c("Agree", "Neutral", "Disagree")))
levels(survey$rating)   #<1>
```

1.  Default is alphabetical — probably not what you want!

✅ Fix with `levels =`:

```{r}
survey$rating <- factor(survey$rating,
  levels = c("Disagree", "Neutral", "Agree"))
levels(survey$rating)
```
:::
:::::::

------------------------------------------------------------------------

## [7. Plotting errors]{style="color:#459B99;"}

::::: columns
::: {.column width="50%"}
**Not enough colors for factor levels**

```         
Error: Insufficient values in manual scale. 
3 needed but only 2 provided.
```

```{r}
#| error: true
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_histogram() +
  scale_fill_manual(values = c("orange", "purple"))   #<1>
```

1.  `iris` has 3 species, but only 2 colors were provided

✅ Fix — provide one color per level (check if `NA` became a factor level too!):

```{r}
#| eval: false
scale_fill_manual(values = c("orange", "purple", "green"))
```
:::

::: {.column width="50%"}
**Extra `+` at the end of a ggplot chain**

```         
Can't add x object
```

```{r}
#| error: true
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram() +   #<1>
```

1.  Trailing `+` with nothing after it

✅ Fix — delete the final `+`

**Extra `%>%` or `|>` in a dplyr chain**

Similarly, re-arranging piped steps and leaving a stray pipe at the end (or middle) will break your code:

```{r}
#| error: true
iris |>
  filter(Sepal.Length > 5) |>   #<1>
```

1.  Dangling pipe — R waits for the next step but there isn't one
:::
:::::

------------------------------------------------------------------------

## [8. R Markdown / Quarto errors]{style="color:#C83532;"}

::::::: columns
::: {.column width="50%"}
**Wrong chunk option names**

```         
Error in options[[sprintf("fig.%s", i)]]
```

Check that your chunk options use the correct names:

| ❌ Wrong         | ✅ Correct           |
|------------------|----------------------|
| `fig.width = 5`  | `out.width = "5in"`  |
| `fig.height = 3` | `out.height = "3in"` |

**Duplicate chunk labels**

```         
Error: Duplicate chunk label 'my-plot'
```

Every code chunk label in your document must be unique. Labels also cannot contain spaces.

```{{r}}
#| label: my-plot-1   #<1>
```

 1. Add a number or descriptor to make each label unique
:::

::::: {.column width="50%"}
**YAML indentation errors**

``` yaml
---
format: html
  toc: true   # ❌ wrong indentation
---
```

✅ Fix:

``` yaml
---
format: 
  html:
    toc: true   # ✅ correct nesting
---
```

:::: red
::: red-cont
**Chunk runs in console but fails on render**

This usually means your document is relying on objects created earlier in your **R session** that don't exist when rendering. The render starts with a **fresh environment**.

Always use **Session → Restart R → Render** to catch these before submitting.
:::
::::
:::::
:::::::

------------------------------------------------------------------------

## Wrap-up: A troubleshooting checklist

When your code gives an error, work through this list:

:::::: columns
::: {.column width="10%"}
:::

::: {.column width="80%"}
| Step | What to do |
|----------------|--------------------------------------------------------|
| **Read** | Read the error message — what function? What keyword? |
| **Typos** | Check spelling, capitalization, commas, and matching parentheses |
| **Packages** | Is the right package installed *and* loaded? Any masking conflicts? |
| **Objects** | Does the variable exist? Run `ls()` to list everything in your environment |
| **Arguments** | Run `?function_name` and check argument names and types |
| **Restart** | Session → Restart R → rerun from top |
| **Google** | Search the exact error message + "in R" |
| **Ask** | Make a reprex and ask a classmate, in office hours, or on Posit Community |
:::

::: {.column width="10%"}
:::
::::::

:::: blue
::: blue-cont
Even experienced programmers encounter errors constantly. Getting faster at *reading* error messages is one of the most valuable skills you will build in this course.
:::
::::

## Typo errors: Incomplete commands

::::: columns
::: {.column width="50%"}
- When the console is waiting for a new command, the prompt line begins with `>`
  - If the console prompt is `+`, then a previous command is incomplete
  - You can finish typing the command in the console window
:::

::: {.column width="50%"}
Example:

```{r}
#| error: true
#| message: true
#| warning: true
#| prompt: true
3 + (2*6
```

```{r}
#| error: true
#| message: true
#| warning: true
#| prompt: true
3 + (2*6
)
```
:::
:::::

## Error Type 1: Typo Errors

**The most common error — and the easiest to fix!**

:::::: columns
::: {.column width="50%"}
**Misspelled function or variable name:**

```{r}
#| error: true
# Typo in function name
my_data <- data.frame(x = 1:5, y = 6:10)
sumamry(my_data)
```

```{r}
#| error: true
# Typo in variable name
my_datta |> head()
```
:::

:::: {.column width="50%"}
**Mismatched parentheses or brackets:**

```{r}
#| error: true
mean(c(1, 2, 3)
```

::: callout-tip
## Quick Fixes

- R is **case-sensitive**: `Mean()` ≠ `mean()`
- Use RStudio's auto-complete (Tab key) to avoid typos
- Check that every `(` has a matching `)`
:::
::::
::::::

## Object is not found

- This happens when text is entered for a non-existent variable (object)

Example:

```{r}
#| error: true
hello
```

- Can be due to missing quotes

```{r}
#| error: true
install.packages(dplyr) 
# correct code is: install.packages("dplyr")
```
