---
title: "Data transformation: Basics"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: Basics"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readxl)
```

```{r}
# library(readxl)
# note we are saving the data from the Excel file in our environment 
# with the object name hrs_data
hrs_data <- read_excel(here("./data/hrs_data_narch.xlsx"))

names(hrs_data)

# Clean up the variable names for ease of use (makes everything lower case here)
hrs_data <- clean_names(hrs_data) # need the janitor package loaded to use this
names(hrs_data)
```

## `dplyr` basics

- `dplyr` is a package in the tidyverse that provides a set of functions for data manipulation and transformation
- Each function in `dplyr` serves almost like a **verb** that describes the action being performed on the data
- Most functions within `dplyr` have similar input and output:
  - Input: a data frame (usually the first argument in the function)
  - Output: a data frame
  - Additional arguments: typically describe what columns (variables) to operate on

## Example of a `dplyr` function: `select()`

- I can use the `select()` function to select specific columns from my data frame

```{r}
hrs_data |> # <1> 
  select( # <2>
    age, # <3>
    ed  # <3>
    ) # <2>
```

1. The first argument is the data frame we want to operate on (in this case, `hrs_data`)
2. `select` is the function we are using on the data frame
3. `age` and `ed` are the columns (in the data frame) that we perform the `select()` function on

## Types of `dplyr` functions

`dplyr` functions can be categorized into four groups based on what they operate on:

1. **rows**: functions that operate on rows of a data frame (e.g., `filter()`, `slice()`)
2. **columns**: functions that operate on columns of a data frame (e.g., `select()`, `mutate()`)
3. **groups**: functions that operate on groups of rows in a data frame (e.g., `group_by()`, `summarize()`)
4. **joins**: functions that combine two data frames based on a common key (e.g., `left_join()`, `inner_join()`)
    - Not covering in this class (You can take BSTA 526 to learn about joins)



