---
title: "Lab 3 Instructions"
author: "Your Name Here"
subtitle: "PUBH 523/623: Data Summarization"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    embed-resources: true
    number-sections: true
    number-depth: 4
    code-annotations: hover
    page-layout: full
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto 
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(gtsummary)
library(here)

# Load your transformed project data from Lab 2

```

## Introduction

With your custom metrics calculated, this lab guides you through quantitative and categorical data summarization using `group_by()`, `summarize()`, and specialized public health descriptive tables using `gtsummary`.

## Lab activities

### Categorical Frequency Checks

Before summarizing data patterns, we need to inspect the distribution of our newly created variables to ensure there are no unintended structural artifacts or empty categories.

::: callout-important

#### Task

Use the `count()` function to generate a frequency distribution table for your asset-flipped outcome variable and your main demographic group. Ensure any `NA` or "Unknown" values are accounted for.
:::

### Stratified Summarization Pipelines

We want to observe how your combined safety net score or access metric varies across groups. You will combine `group_by()` and `summarize()` to extract key descriptive summary statistics.

::: callout-important

#### Task

Write a tidyverse pipeline that groups your data by your demographic factor **and** your asset-flipped outcome variable. Use `summarize()` to calculate:

1. The total count (`n()`) of households per cell.
2. The mean of your combined safety net metric.
3. The standard deviation or median of your metric.
Make sure to use `na.rm = TRUE` to handle any missing values safely.
:::

### Building a Professional Table 1

In public health reporting, a structured "Table 1" describes the characteristics of your study population. We will utilize the `gtsummary` package to build a beautiful, publication-ready table.

```{r}
#| eval: false
#| label: table1-template

# Example template layout
wbns_cleaned %>%
  select(demographic_var, asset_metric, outcome_var) %>%
  tbl_summary(
    by = outcome_var,
    label = list(demographic_var ~ "Race/Ethnicity", asset_metric ~ "Active Support Programs")
  ) %>%
  add_p() %>%
  bold_labels()

```

::: callout-important

#### Task

Generate a descriptive `tbl_summary()` object stratified by your asset outcome variable. Customize your variable labels so they show up as clean, professional English titles rather than shorthand code names (e.g., change `PPETHM` to "Race and Ethnicity").
:::
