---
title: "Data transformation: *Advanced*"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: *Advanced*"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  eval: false
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio
)
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_data <- clean_names(hrs_data_rds) |>
  relocate(race_original, sex, .after = degree) |>
  relocate(age_yr, ed, cesd, .after = birthdate) # |>
  # select(id, age_yr, ed, cesd, degree, srh)
```

# Rows

## Filtering requires a little logic

-   We can chain multiple criteria using the [`&` (AND)]{style="color:darkorange"} or [`|` (OR) operators]{style="color:darkorange"}.
    -   But we need to review a little logic before we do this.



-   If we wanted to return participants who identified as
    -   `Female` **and** are in 10th grade,
    -   we would use an `&` to chain these criteria together:

```         
sex == "Female" & grade == "10th"
```



-   Note that *SPELLING AND CASE MATTER*!
    -   You must write the name of the column exactly and
    -   also the value (i.e. "Female" not "female") exactly.

## Filtering with `&` (AND) criteria

```{r}
hrs_data |>
  filter(diab == "Yes" & 
           degree == "Master's degree")

```

-   Note that we could also use the comma in this case:

```{r}
hrs_data |>
  filter(diab == "Yes",    # <1> Comma is equivalent to "&" within filter()
         degree == "Master's degree")

```

## Filtering with `|` (OR) criteria

If we wanted patients who were

`Female` [**or**]{style="color:darkorange"} in 10th grade

we would use a `|` to chain these criteria together.

```{r}
hrs_data |>
  filter(diab == "Yes" |
           degree == "Master's degree")
```

::: {.callout-warning icon="false"}
## Think about it:

Which of the AND vs. OR code blocks will return a larger number of participants?
:::

## More filter examples

-   See [this slide](https://jminnier-berd-r-courses.netlify.app/02-data-wrangling-tidyverse/02_data_wrangling_slides_part1.html#27){target="_blank"} from a previous BERD workshop.

-   Also read the [filter section](https://r4ds.hadley.nz/data-transform.html#rows){target="_blank"} in the "Data Transformation" chapter from *R for Data Science (2e)*.

<br>

### More about comparison and logical operators

Click [here](https://www.datamentor.io/r-programming/operator/){target="_blank"} for a useful reference for all the different operators (both logical and comparison) that you can use.

## 

Using %in%

# `arrange()`

## `arrange()` after `count()`-ing categorical data

-   Another example of when `arrange()` is useful

::: columns
::: {.column width="48%"}
-   We can also use `count()` to count all of the unique values for a categorical variable.
-   This will also show `NA` category, which is the special missing variable type:

```{r}
hrs_data |>
  count(age)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
-   Sometimes it helps to arrange by that category if you have many categories to view.
-   This can also alert us if there are categorical coding mistakes (such as misspellings) in our data.

```{r}
hrs_data |>
  count(age) |>
  arrange(n)
```
:::
:::

# Columns

## Categorical data (`factor`s) 

::: {.panel-tabset}

### Intro

-   One data type that we haven't yet looked at are `factor` variables
    -   [factors are how R represents **levels** of *categorical data*]{style="color:darkorange"}.
-   For the most part, you can use `character` and `factors` interchangeably for categorical data.
-   However, there are two main differences:
    -   `factor`s define the *permissible* values in a vector with an argument called `levels`.
    -   The levels define the order in which the permissible values are displayed, such as in a table or figure.
    -   The default order of levels (values) is in **alphanumeric** order.



### Make `degree` factor

-   Below we use the `factor()` function within a `mutate()` to transform `degree` from a `character` variable into a `factor` variable.

```{r}
class(hrs_new$degree) # what's the data type of degree?
hrs_new |> tabyl(degree)

hrs_new <- hrs_new |> 
  #create a factor variable
  mutate(degree_fac1 = factor(degree)) 

# Checks:
class(hrs_new$degree_fac1) # what's the data type of degree_fac1?
levels(hrs_new$degree_fac1) # what are its levels?
# values are the same in the two degree variables:
hrs_new |> tabyl(degree, degree_fac1) |> adorn_title()
```

### Change order of levels

-   Use the `levels` argument within the `factor()` function

Example: Create `degree_fac2` variable as a factor version of degree with degrees in "correct" order:

```{r}
hrs_new <- hrs_new |> 
  #create a factor variable
  mutate(degree_fac2 = factor(
    degree, 
    levels = c("9th","10th","11th","12th")
    ))

# Checks:
class(hrs_new$degree_fac2) # what's the data type of degree_fac1?
levels(hrs_new$degree_fac2) # what are its levels?
hrs_new |> tabyl(degree_fac2)

# values are the same in the two degree factor variables:
hrs_new |> tabyl(degree_fac1, degree_fac2) |> adorn_title()
```

### Be careful! 

-   Recall that the `levels` of a factor are the *permissible values* in a `factor`.

::: callout-warning
*Spelling matters!*
:::

-   If we try to use a factor level that doesn't exist ("10" vs "10th"),
    -   we won't have any data in that category, and
    -   the non-specified levels ("10") get set to `NA` (missing).

```{r}
hrs_new <- hrs_new |> 
  mutate(degree_fac3 = factor(
    degree, 
    levels = c("9th","10","11th","12th") # typo!!
    )) 

# check
levels(hrs_new$degree_fac3) # what are its levels?
hrs_new |> tabyl(degree_fac3)
# compare
hrs_new |> tabyl(degree_fac2, degree_fac3) |> adorn_title()
```


### Replace `degree` with factor version

-   Below we are doing something called *reassignment*:
    -   We’re taking the previous values of our variable (`degree`),
    -   doing something to them (making them a factor), and 
    -   then reassigning the variable `degree` to its new set of values.
-   *I don't recommend replacing a variable with new values until you are 100% positive that your code is correct!!*
    -   I can't compare the new `degree` values to the original ones anymore to check my work

```{r}
hrs_new <- hrs_new |> 
  mutate(degree = factor(
    degree, 
    levels = c("9th","10th","11th","12th")
    )) 

# compare to degree_fac2 (which we determined to be correct)
hrs_new |> tabyl(degree, degree_fac2) |> adorn_title()
```

### Remove extra variables

-   This step isn't necessary, but we don't need the columns `degree_fac1`, `degree_fac1-degree_fac3` anymore.

```{r}
names(hrs_new)
hrs_new <- hrs_new |> select(-degree_fac1:-degree_fac3)
names(hrs_new)
```


:::



# Practice time! (part 5)

1.  Create a new variable `income` that calculates income based on `income` and `weight_kg`. Note the formula for income is weight/income\^2 (weight is in kilograms and income is in meters). Create a scatterplot of `income` vs `income`.

2.  Create a new factor variable `female_fac` that orders the values of `female` as Male, Female.

# Make variables with `case_when()`

![[Artwork by \@allison_horst](https://allisonhorst.com/everything-else){target="_blank"}](image/dplyr_case_when.png){fig-align="center"}

## `case_when()`

::: {.panel-tabset}

### Intro

-   How can we dichotomize a continuous variable, such as income?
-   How can we combine categories of a variable, such as race?
-   How can we convert categories to numbers, such as female?

These are all examples where `case_when()` can be used.

### Code

-   The code follows a **basic pattern** for each category inside `case_when()`:
    -   [*condition* \~ *category name*]{style="color:darkorange"}
-   The *left* side of the `~` is where we specify how to *define* the category.
-   The *right* side of the `~` is where we specify the *category name*.
-   Separate each "case" with a comma.

### Example 1

```{r}
hrs_new |> tabyl(female)
```


-   Convert `female` to a new numeric variable called `female` whose values are
    -   1 when `female` is Female, and
    -   0 when `female` is Male

[*left side* defines the condition *\~* *right side* names the category]{style="color:darkorange"}

```{r}
hrs_new <- hrs_new |>
  mutate(
    female = case_when(
      # when female is Female, make female equal to 1 
      female == "Female" ~ 1, # do not forget this comma!
      
      # when female is Male, make female equal to 0
      female == "Male" ~ 0
      )
    ) 

# checks
hrs_new |> tabyl(female) # look at the categories
class(hrs_new$female)
# compare female and female variables
# hrs_new |> tabyl(female, female) |> adorn_title()
```

-   How were the missing values (`NA`) handled?



### Example 2

-   Dichotomize the numeric variable `income` into two categories with 30 as the cutoff.

[*left side* defines the condition *\~* *right side* names the category]{style="color:darkorange"}

```{r}
hrs_new <- hrs_new |>
  mutate(
    income30 = case_when(
      income < 30 ~ "<30", # don't forget a comma between cases
      income >= 30 ~ "30+"
      ) 
  ) 

# check
hrs_new |> tabyl(income30)
class(hrs_new$income30)

# check values were categorized correctly: (look at mins & maxes)
hrs_new |> 
  group_by(income30) |> 
  get_summary_stats(income) |> 
  gt()

```


### Example 3

-   Recode `income` to have 4 levels:
    -   `<20` (not including exactly 20)
    -   `20-30` (not including exactly 30)
    -   `30+`
    -   `unknown` if missing

-   Code comments
    -   Notice that for the middle category we use an `&` statement similar to how we used logic in filter
    -   New: used `.default` option to specify what to assign in all other unspecified cases.

```{r }
hrs_new <- hrs_new |>
  mutate(
    income3cat = case_when(
      income < 20 ~ "<20",
      (income >=20) & (income < 30) ~ "20-30",
      income >= 30 ~ "30+",
      .default = "unknown"  # apply this in all other cases
      ),
    income3cat = factor(income3cat) # make it factor after creating categories
    )
  
# check
hrs_new |> tabyl(income3cat)
class(hrs_new$income3cat)

# check values were categorized correctly: (look at mins & maxes)
hrs_new |> 
  group_by(income3cat) |> 
  get_summary_stats(income) |> 
  gt()
```

:::


# Replace missing values with `replace_na()`

## `replace_na()`

-   Sometimes we want to fill in missing values with a certain value.
-   We can use the `replace_na()` function inside of `mutate()` to specify this.

::: {.panel-tabset}

### Example 1

-   For example, suppose if the income variable is missing,
    -   we want to replace it with the average income across the entire data set.

```{r}
hrs_new <- hrs_new |>
  mutate(
    income_filled = replace_na(
      income,
      mean(income, na.rm = TRUE))
    )           

# check
hrs_new |> 
  select(contains("income")) |> 
  tail()
# how does income_filled compare to income?
hrs_new |> 
  select(income, income_filled) |> 
  get_summary_stats() |> 
  select(-iqr, -se, -ci) |> 
  gt()

```

### Example 2

-   We can also fill in missing values for character variables
-   Here we replace missingness in the `female` column with `unknown`:

```{r}
class(hrs_new$female)
hrs_new |> tabyl(female)

hrs_new <- hrs_new |>
    mutate(
      femalefilled = replace_na(female, "unknown"))

# check
hrs_new |> tabyl(femalefilled, female) |> adorn_title()
```

### Example 3

-   What happens when we use `replace_na()` to fill in missing values for factor variables?

```{r}
class(hrs_new$degree)
```


```{r}
#| error: true

hrs_new <- hrs_new |>
    mutate(
      degreefilled = replace_na(degree, "unknown"))
```

:::

# `select()` 

## `{tidyselect}`

-   There are ways to select columns by "searching" for them.
-   These are called `{tidyselect}` helpers.
-   See more examples from the [{tidyselect} webpage](https://tidyselect.r-lib.org/reference/language.html){target="_blank"}.


### `contains`

-   For instance, you can see how this might be useful, where we select all columns where the column name includes the word "race":

```{r}
hrs_data |> 
  select(contains("race"))
```

### `starts_with`, `ends_with`

-   Select all columns starting with "race":

```{r}
hrs_data |> 
  select(starts_with("prob")) |> 
  names()
```

-   Select all columns ending with "e":

```{r}
hrs_data |> 
  select(ends_with("e")) |> 
  names()
```

### Select columns by data type

-   We can also select variables based on data types
-   For instance, `select(where(is.numeric))` selects all numeric columns
    -   (`where()` is a helper function that returns columns where the inside function is `TRUE`,
    -   in this case `is.numeric` returns the `TRUE`/`FALSE`)


```{r}
hrs_data |> 
  select(where(is.numeric)) |> 
  names()

hrs_data |> 
  select(where(is.character)) |> 
  names()

```


