---
title: "Piping"
author: "Nicky Wakim, Jessica Minnier, Meike Niederhausen"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Piping"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  janitor,
  here
)
```

```{r}
#| echo: false
hrs_data <- read_rds(here("./data/hrs_data.rds"))
hrs_data <- clean_names(hrs_data) |>
  relocate(age_yr, .after = birthdate) # move age_yr column to be after id column
```

## What is piping?

- Pipes (and piping) are used to pass the output of one function as input into another function
- You can think of pipes as "then" statements

::: blue
::: blue-cont
I want to find my keys, then start my car, then drive to work, then park my car.
:::
:::

::: columns
::: {.column width="50%"}
**Nested**

```{r eval = FALSE}
park( #<4>
  drive( #<3>
    start_car( #<2>
      find("keys") #<1>
      ),  #<2>
    to = "work"
    ) #<3>
  ) #<4>
```

1. Our first action is embedded the furthest into the nested functions
2. Then the output of "find keys" in the input of "start car" which is nested as well
3. We keep working our way out to "drive"
4. Then we finally park
:::

::: {.column width="50%"}
**Piped**

```{r eval = FALSE}
find("keys") |> #<1>
  start_car() |> #<2>
  drive(to = "work") |> #<3>
  park() #<4>
```

1. First statement is the first thing we want to do
2. Output of "find keys" is inputted into "start car"
3. Then we drive to work
4. Then we park the car

- Pipes tend to match our sequential thoughts better
:::
:::

## Two types of pipes you might see

- Two types: `|>` vs `%>%`
  - `%>%` is the original pipe in R (requires `magrittr` package within `tidyverse`)
  - `|>` is a newer pipe (started in 2021 and does not require a package)
- Both pipes are used to write code from **left to right** 
    -   by passing the result of one step into the next

::: {.columns}
                                        
::: {.column width="50%"}

::: red
::: red-ttl
Example using `|>`
:::
::: red-cont
```{r}
hrs_data |> dim()
```
:::
:::

:::
                                          
::: {.column width="50%"}
::: blue
::: blue-ttl
Example using `%>%`
:::
::: blue-cont
```{r}
hrs_data %>% dim()
```
:::
:::
:::
:::

- You can use either! Our notes will use `|>` since it is generally where the field is headed.

## Simple vs complicated input

::: pink2
::: pink2-cont
[**Simple:** when the receiving function only has one input argument, we can use:]{style="color:#BF396F;"}
:::
:::

::: columns
::: {.column width="50%"}
```{r}
hrs_data |> dim() #<1>
```

1. Just the function name `dim` with parentheses.
:::
::: {.column width="50%"}
```{r}
hrs_data %>% dim() #<1>
```
1. Just the function name `dim` with parentheses.

:::
:::

 

::: green
::: green-cont
[**Complicated:** when the receiving function has more input arguments, we need to specify **where the piped input goes**. We do this with a placeholder:]{style="color:#459B99;"}
:::
:::
::: columns
::: {.column width="52%"}
The placeholder for `|>` is `_`
```{r}
#| eval: false
hrs_data |> #<1>
  lm(height ~ age_yr, data = _) #<2>
```

1. We take the `hrs_data` and pipe it into a function
2. This function `lm` requires multiple inputs. `height ~ age_yr` is one input and `hrs_data` is the other. `hrs_data` is represented by `_` since it is piped through. 
:::
::: {.column width="48%"}
The placeholder for `%>%` is `.`
```{r}
#| eval: false
hrs_data %>% #<1>
  lm(height ~ age_yr, data = .) #<2>
```

1. We take the `hrs_data` and pipe it into a function
2. This function `lm` requires multiple inputs. `height ~ age_yr` is one input and `hrs_data` is the other. `hrs_data` is represented by `.` since it is piped through. 
:::
:::

## Notice the code styling

- We can start a new line after every pipe (`|>`)
  - R reads this as an incomplete line and will continue to read

```{r}
#| eval: false
hrs_data |> #<1>
  lm(height ~ age_yr, data = _) #<2>
```

1. With the `|>` R will look to the next line for where `hrs_data` is inputted
2. Then we follow with our desired function

::: columns
::: column
- If you have many functions in a row, putting each on a new line makes your code easy to follow!

```{r}
#| eval: false
hrs_data |> #<1>
  select(height, age_yr) |> #<2>
  lm(height ~ age_yr, data = _) #<3>
```

1. With the `|>` R will look to the next line for where `hrs_data` is inputted
2. Then I select certain columns of my dataset
3. Then I run that through the `lm` function
:::
::: column
- You can further style the functions to make their input readable

```{r}
#| eval: false
hrs_data |> #<1>
  select( #<2>
    height, 
    age_yr
    ) |> #<2>
  lm( #<3>
    height ~ age_yr, 
    data = _
    ) #<3>
```

1. With the `|>` R will look to the next line for where `hrs_data` is inputted
2. `select` is not styled to highlight the input
3. `lm` is now styled to highlight the input
:::
:::

## Resources

- [Tidyverse style guide on pipes](https://style.tidyverse.org/pipes.html)
- More on the two pipes
  -   [Differences between the base R and magrittr pipes](https://tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/){target="_blank"} - Tidyverse blog post by Hadley Wickham
  - [Discussion of differences on Stack Overflow](https://stackoverflow.com/questions/67633022/what-are-the-differences-between-rs-native-pipe-and-the-magrittr-pipe){target="_blank"}
  - [R 4.1.0 pipe overview](https://www.jumpingrivers.com/blog/new-features-r410-pipe-anonymous-functions/){target="_blank"}
  -   [Comparing pipes: Base-R |> vs {magrittr} %>%](https://albert-rapp.de/posts/31_pipes_compared/31_pipes_compared){target="_blank"} 
      -   this resource includes examples and 
      -   also how to change the "default" pipe in RStudio from `%>%` to `|>` (keyboard shortcut will use `|>`)