---
title: "Functions"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Functions
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  janitor,
  here
)
```

## What are functions?

- R has many built-in functions that allow you to perform a variety of tasks

- Functions usually take the form:

```{r}
#| eval: false

function_name(                           # <1>
  argument1 = input_value1,              # <2>
  argument2 = input_value2,              # <2>
  ...                                    # <2>
  )                                      # <1>
```

1. `function_name` is the name of the function we are using.
2. Within the parentheses of the function, there are set arguments that we assign values to. 

## Example: simple function like `mean()`

- `mean()` is a function that generates the calculated mean of a set of numbers
- It has one main argument: `x`
- Output (automatically printed) will be the mean of the numbers in `x`

```{r}
mean(
  x = c(1, 2, 3, 4, 5, 6) #<1>
)
```

1. The argument `x` is assigned the value of a vector of numbers (1, 2, 3, 4, 5)

## Example: calling the `seq()` function

- `seq()` is a function that generates a sequence of numbers. 
- It has three arguments: `from`, `to`, and `by`
- Output will be the sequence numbers 

```{r}
seq(
  from = 1, #<1>
  to = 10, #<2>
  by = 2 #<3>
)
```

1. The argument `from` is assigned the value of 1, which is the starting point of the sequence.
2. The argument `to` is assigned the value of 10, which is the ending point of the sequence.
3. The argument `by` is assigned the value of 2, which specifies the increment between numbers in the sequence.

## We don't need to specify the argument names

- We can run `seq()` without the argument names
- R will assume that the values are in the correct order of the arguments

 

We get the same result:

```{r}
seq(1, 10, 2)
```

## Output of a function

- We can also assign the output of a function to an object using the assignment operator `<-`:

```{r}
#| eval: false

function_output <- function_name(        # <1>
  argument1 = input_value1,              # <2>
  argument2 = input_value2,              # <2>
  ...                                    # <2>
  )                                     
```

1. The output of `function_name` is assigned to an object called `function_output`
2. Within the parentheses of the function, there are set arguments that we assign values to. 


- The output will not print, but it will be stored in the object `function_output`

## Now with the `seq()` function

- We will assign the output of the `seq()` function to an object called `odds`:

```{r}
odds <- seq(
  from = 1, 
  to = 10, 
  by = 2
)
```

- So `odds` will now contain the sequence (1,3,5,7,9)

## Help with functions

- If you need a reminder of how to use a function, you can use the `?` or `??` to access the help file for that function:
```{r}
#| eval: false
?seq #<1>
??read_sas #<2>
```

1. The `?` will take you to the help file for the `seq()` function, which will show you the arguments and how to use it.
2. The `??` will search for any functions `read_sas` from your *installed* packages and show you a list of functions that match that search term. This is helpful when you don't remember the package a function is in.

 

- Certain functions will show you the arguments if you press `Tab` after typing the function name and parentheses
```{r}
#| eval: false
mean()
sd()
```

 

- R will also help autofill the names of functions from your loaded packages:
```{r}
#| eval: false
m
s
```

