---
title: "Common errors"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Common errors"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  # janitor,
  here, 
  rio
)
```

## Errors and troubleshooting are a huge part of coding!


<style>
.small-error pre,
.small-error .cell-output-error {
  font-size: 0.55em;
}
</style>

:::: pink
::: pink-cont
A huge part of coding is **learning to debug your own code** and **figuring out what's going wrong.**
:::
::::

- Errors can feel frustrating at first, but they are R trying to *help* you.

We will cover:

1.  [How to interpret error messages]{style="color:#BF396F;"}
2.  [The most common types of errors and how to fix them]{style="color:#3070BF;"}

## Error vs. Warning vs. Message


| Type        | Meaning                                          |
|-------------|--------------------------------------------------|
| **Error**   | Code stopped; something went wrong               |
| **Warning** | Code ran, but R is flagging something suspicious |
| **Message** | Informational; not a problem                     |

 

Example of a warning:
```{r}
#| warning: true

log(-1)  
```

- In this case, the code ran, but R is warning you that the result is `NaN` (not a number) because the logarithm of a negative number is undefined

## Interpreting error messages

- R error messages can feel cryptic, but they usually follow a pattern

::::: blue
::: blue-ttl
Pay attention to
:::

::: blue-cont
- Which **function** is mentioned in the error
- **Keywords** like `not found`, `unexpected`, `unused argument`, or `subscript out of bounds`
- Any **line numbers** or column names that appear
:::
:::::

## Types of common errors

We will go through the following error messages:

- Misspelled function or variable name
- Mismatched parentheses
- Function not found
- Object not found
- File not found
- Subscript out of bounds
- Wrong argument name
- Silent logic errors


:::: pink
::: pink-cont
We will go through each type with real error messages and how to fix them. Keep this slide as a reference!
:::
::::

## Misspelled function or variable name

::: blue
::: blue-ttl
R is case-sensitive
:::
::: blue-cont
`Mean()` and `mean()` are different things!

```{r}
#| error: true
Mean(c(1, 2, 3))   #<1>
```

1.  Capital M — R can't find this function

```{r}
mean(c(1, 2, 3)) 
```
:::
:::


::: pink
::: pink-ttl
Accidental mispelling
:::
::: pink-cont
```{r}
#| error: true
my_data <- data.frame(x = 1:5, y = 6:10)
sumamry(my_data)   #<2>
```

2.  Typo in the function name (`sumamry` instead of `summary`)
:::
:::

## Mismatched parentheses

::: {.small-error}
```{r}
#| error: true
mean(c(1, 2, 3)   #<1>
```
:::

- Missing closing `)` — the console shows `+` waiting for more input

:::: blue
::: blue-cont
When the console prompt is `>`, R is ready for a new command. When it switches to `+`, a previous command is **incomplete** — finish typing it, or press **Escape** to cancel and start over.
:::
::::

**Quick fixes:**

- Use **Tab** to autocomplete function names and avoid typos
- Check that every `(` has a closing `)`
- Use RStudio's bracket highlighting — click next to a `(` and its match will highlight

## Function not found, package not loaded

- This usually means you typed the function name incorrectly, or **forgot to install or load the package**

```{r}
#| error: true
clean_names(iris)   #<1>
```

1.  `clean_names()` is from `janitor`: if it's not loaded, R can't find it

::: orange
::: orange-ttl
Fix: make sure to load the package!
:::
::: orange-cont

```{r}
pacman::p_load(janitor)      #<2>
iris_clean <- clean_names(iris)
names(iris_clean)
```

2.  Load the package first, then call the function

:::
:::

::: pink
::: pink-ttl
You can also use `::` to call it without loading the whole package
:::
::: pink-cont
```{r}
iris_clean2 <- janitor::clean_names(iris)   #<3>
names(iris_clean2)
```

3.  Recall from the Packages lesson: `package::function()` lets you use one function without `library()`

:::
:::


## Object not found

This means R can't find a variable you referenced. Common causes:

[1. You never created the object]{style="color:#BF396F;"}

- This can happen if you ran something in the console, but don't have it in your current `.qmd` file when rendering

```{r}
#| error: true
my_summary   #<1>
```

1.  This object was never created: code above may not have run

 

[2. You're missing quotes around a package or a string]{style="color:#459B99;"}
```{r}
#| error: true
install.packages(dplyr)   #<2>
```

2.  Missing quotes: R looks for an object named `dplyr`, not the text "dplyr"

    - Fix: `install.packages("dplyr")`

## File not found

- This happens when R can't locate a file you're trying to read in

```{r}
#| error: true
import(here("data/hrs_Data.csv"))   #<2>
```

 

For now, if you run into it, check:

- Spelling of the file name
- Forward slashes (MAC) vs. backward slashes (Windows)
- Correct file extension (`.csv`, `.xlsx`, `.rds`)
- That you're working inside the right **R Project**, so your working directory is where you think it is 

## Subscript out of bounds

- This happens when you try to access an **element** of a vector, list, or data frame that doesn't exist

```{r}
#| error: true
x <- c(1, 2, 3, 4, 5)  #<1>
x[3:10]  #<2>
```

1.  `x` is a vector of length 5
2. I try to access the 3rd to 10th elements, but some of them don't exist.

- There is no error, but R returns `NA` for the elements that don't exist

 

- Here's an example of the error wording (if an error is thrown):

```{r}
#| error: true
x[[10]]  #<3>
```

3. I try to access the 10th element iwth double brackets, but it doesn't exist.

## Wrong argument name

- This can happen if you misremember the argument name, or if you accidentally type it incorrectly

::::::: columns
::: {.column width="60%"}
```{r}
#| error: true
mean(c(1, 2, NA), remove.na = TRUE)   #<1>
```

1.  `remove.na` is not a real argument: R silently ignores it, and `NA` propagates through

:::: pink
::: pink-cont
**Notice this one didn't throw an error!** It just gave you the wrong answer! Not every mistake is loud. 
:::
::::

- Fix: use the correct argument name

```{r}
mean(c(1, 2, NA), na.rm = TRUE)   #<2>
```

2.  The correct argument is `na.rm`
:::

::: {.column width="5%"}
:::

::: {.column width="35%"}
::::: orange
::: orange-ttl
How to find the right argument names
:::

::: orange-cont
Run `?function_name` and read the **Arguments** section — it lists every accepted argument, what it expects, and its default value.

```{r}
#| eval: false
?mean
```
:::
:::::

:::
:::::::

## Silent logic errors

- This happens when your code runs without throwing an error, but the 'TRUE'/'FALSE' output is not what you expected

```{r}
0.1 + 0.2 == 0.3   #<1>
```

1.  You'd expect `TRUE`, but computers store decimals in binary, and `0.1 + 0.2` is actually stored as a number *very slightly* off from `0.3`, so the **comparison quietly returns `FALSE`**

 

- Fix: use `all.equal()` or a small tolerance instead of `==` when comparing decimals:

```{r}
all.equal(0.1 + 0.2, 0.3)  #<1>
```

1.  `all.equal()` checks whether two numbers are "close enough," rather than checking for exact equality

:::: pink
::: pink-cont
Logic errors are the sneakiest bugs. Always **check your output** or check an example if you're running many calculations. 
:::
::::

## Wrap-up

- Errors are a normal part of coding! 
  - Don't be discouraged, R is trying to help you
- When you get an error, read it carefully and try to understand what it's telling you
- Recognize the most common types of errors and what might have caused them


