---
title: "Common Errors"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Common Errors"
    html-math-method: mathjax
    highlight-style: arrow
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

## Errors and troubleshooting are a huge part of coding!

## Types of common errors

1. Typo errors
2. Package errors
3. Object errors
4. Function syntax errors
5. Factor errors
6. R Markdown errors

## Incomplete commands

::: columns
::: {.column width="50%"}
-   When the console is waiting for a new command, the prompt line begins with `>`
    -   If the console prompt is `+`, then a previous command is incomplete
    -   You can finish typing the command in the console window
:::

::: {.column width="50%"}
Example:

```{r}
#| error: true
#| message: true
#| warning: true
#| prompt: true
3 + (2*6
```

```{r}
#| error: true
#| message: true
#| warning: true
#| prompt: true
3 + (2*6
)
```
:::
:::

## Object is not found

-   This happens when text is entered for a non-existent variable (object)

Example:

```{r}
#| error: true
hello
```

-   Can be due to missing quotes

```{r}
#| error: true
install.packages(dplyr) 
# correct code is: install.packages("dplyr")
```