---
title: "Workflow: Naming"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Workflow: Basics"
    html-math-method: mathjax
    highlight-style: arrow
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

## Naming

- File naming
- variable naming

## What's in a name? {#sec-whats-in-a-name}

Object names must start with a letter and can only contain letters, numbers, `_`, and `.`.
You want your object names to be descriptive, so you'll need to adopt a convention for multiple words.
We recommend **snake_case**, where you separate lowercase words with `_`.

```{r}
#| eval: false
i_use_snake_case
otherPeopleUseCamelCase
some.people.use.periods
And_aFew.People_RENOUNCEconvention
```

We'll return to names again when we discuss code style in @sec-workflow-style.

You can inspect an object by typing its name:

```{r}
x <- 3
x
```

##

Make another assignment:

```{r}
this_is_a_really_long_name <- 2.5
```

To inspect this object, try out RStudio's completion facility: type "this", press TAB, add characters until you have a unique prefix, then press return.

Let's assume you made a mistake, and that the value of `this_is_a_really_long_name` should be 3.5, not 2.5.
You can use another keyboard shortcut to help you fix it.
For example, you can press ↑ to bring the last command you typed and edit it.
Or, type "this" then press Cmd/Ctrl + ↑ to list all the commands you've typed that start with those letters.
Use the arrow keys to navigate, then press enter to retype the command.

##

Change 2.5 to 3.5 and rerun.

Make yet another assignment:

```{r}
r_rocks <- 2^3
```

##

Let's try to inspect it:

```{r}
#| eval: false
r_rock
#> Error: object 'r_rock' not found
R_rocks
#> Error: object 'R_rocks' not found
```

This illustrates the implied contract between you and R: R will do the tedious computations for you, but in exchange, you must be completely precise in your instructions.
If not, you're likely to get an error that says the object you're looking for was not found.
Typos matter; R can't read your mind and say, "oh, they probably meant `r_rocks` when they typed `r_rock`".
Case matters; similarly, R can't read your mind and say, "oh, they probably meant `r_rocks` when they typed `R_rocks`".

