---
title: "R projects and working directories"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "R projects and working directories"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  lubridate,
  janitor,
  here
)
```

## Why do we need R Projects and the `here` package?

:::::::: columns
::: {.column width="35%"}
:::

::::: {.column width="30%"}
:::: pink
::: pink-cont
**Answer:** Reproducibility!
:::
::::
:::::

::: {.column width="35%"}
:::
::::::::

- Research data and code can reach the same results regardless of who is running the code

  - This can also refer to future or past you!

- We want to set up our work so the entire folder can be moved around and work in its new location

## What is a working directory?

::::: blue
::: blue-ttl
Working directory
:::

::: blue-cont
The **working directory** is the folder R is currently "looking at." When you want to locate a file, it is the starting point for all file paths in your session.
:::
:::::

- The following will show me my current working directory

```{r}
getwd() #<1>
```

1.  Function to look at the current working directory, which is where my `.qmd` file lives

- Any **relative path** starts from the current working directory

```{r}
#| eval: false

read_rds("../data_lessons/hrs_data.rds") #<1>
```

1.  If I want to go up a folder, I can use `..` to go up one level, and then I can access the `data_lessons` folder and the `hrs_data.rds` file.

- Note that the **absolute file path** is `/Users/wakim/Library/CloudStorage/OneDrive-OregonHealth&ScienceUniversity/Teaching/Classes/PUBH_523_26Su/PUBH_523_26Su_site/lessons/` `data_lessons/hrs_data.rds`

## The problem with working directories

- You *can* set the working directory by hand, but you should NOT!

- When you set a `wd` manually, it will break if you move the folder or if you are on a different computer

```{r}
#| eval: false

setwd("/Users/jane/Desktop/my_project") #<1>

setwd("C:/Users/jane/Documents/school/BIOL401/my_project") #<2>
```

1.  Breaks on any other computer
2.  Also breaks if you move the folder

- When you use `qmd`s, running code in the console and rendering result in two different working directories
  - Let's walk through an example!

:::: pink
::: pink-cont
R projects and the `here` package can help **fix both of these issues**!
:::
::::

## What is an R Project?

::::: blue
::: blue-ttl
R Project
:::

::: blue-cont
An **R Project** is a file (`.Rproj`) that sits in a folder and tells RStudio: "This is where I want my working directory to be set automatically."
:::
:::::

:::::::::::: columns
::::: {.column width="30%"}
:::: pink
::: pink-cont
I can add an R Project to my `PUBH_523_26Su` folder. The R project will show up as `PUBH_523_26Su.Rproj`.
:::
::::
:::::

::: {.column width="5%"}
:::

:::: {.column width="35%"}
::: large-code
```         
PUBH_523_26Su/ 
├── PUBH_523_26Su.Rproj
├── admin/
├── data_lessons/
├── data_practice/
├── data_project/
│   ├── data_raw/  
│   └── data_processed/    
├── lessons/
├── practice/
├── project/
└── README.txt
```
:::
::::

::::: {.column width="30%"}
:::: green
::: green-cont
When you open the `.Rproj` file, RStudio sets the working directory to that folder — every time, on any computer.
:::
::::
:::::
::::::::::::

## Let's create an R Project!

- You can create a [*new project and folder*]{style="color:#C83532"}
  - **File → New Project → New Directory → New Project**
  - Give the project a name, choose where to save it, and click **Create Project**

 

- You can also create a project from an [*existing folder*]{style="color:#459B99"}
  - **File → New Project → Existing Directory**
  - Then you'll browse for the folder you want to create the project in

 

Let's see an example!

## Opening and using an R Project

- When working with files in an R Project, you want to open and work within the R project

 

- You can open an R project two ways:
  - Click on the `.Rproj` file in your folder
  - Open RStudio and select the project from the **top-right corner** of the window

 

- Once a project is open, we can work in our files and do not need to set a working directory manually
  - R Studio will automatically set the working directory to the project folder

:::: red
::: red-cont
Opening a script/qmd directly (by double-clicking it) does **not** activate the project or set the working directory
:::
::::

## We still run into some issues with R Projects

We haven't fixed this issue yet:

> - When you use `qmd`s, running code in the console and rendering result in two different working directories
>   - Let's walk through an example!

:::: pink
::: pink-cont
We need the `here` package to fix this issue!
:::
::::

## The `here` package

- For scripts in subfolders, use `here()` to build paths from the project root

- The `here()` function makes your code start at the R project folder whether we are working in the console or rendering a `qmd` document

  - Output for `here()` returns same thing in console and on render

  ```{r}
  pacman::p_load(here) #<1>
  here() #<2>
  ```

  1.  Need to load the `here` package to use the `here()` function
  2.  The `here()` function returns the path to the project root folder, regardless of where the script lives

## Now we can navigate to our files in a robust way

- We can use `here()` to navigate to files in our project folder
- We only need to specify the relative path from the project root folder
- We can input the relative path two different ways:
  - Using a comma to separate the folder/file names (*better practice bc it is platform agnostic*)
  - Using a forward (or backward on Windows) slash to separate the folder/file names

```{r}
here("lessons", "01_Welcome")

here("images", "PUBH_R.png")
here("images/PUBH_R.png")
```

## Wrap-up

::::::: columns
::: {.column width="55%"}
- Create **one R Project** per analysis

 

- Open work by clicking the **`.Rproj` file** or opening in R Studio

 

- Use **relative paths**: no `setwd()`, ever

 

- Use **`here()`** to make relative file paths
:::

::::: {.column width="45%"}
:::: pink
::: pink-cont
R Projects $+$ `here()` $+$ good folder structure

$$=$$

analysis that runs reliably anywhere, for anyone
:::
::::
:::::
:::::::
