---
title: "Getting help"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Getting help"
    html-math-method: mathjax
    highlight-style: arrow
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

## Main ways to get unstuck

1. [Use the `?` in front of the function name to get more information!]{style="color:#BF396F;"}
  
    -   *Best for: understanding function arguments and usage*

2. [Google or go to stackoverflow.com]{style="color:#3070BF;"}
  
    - *Best for: specific errors and edge cases*

3. [I can also go to my favorite AI tool to get help]{style="color:#459B99;"}
  
    - *Best for: getting code started or explaining errors*

4. [Ask another person!]{style="color:#C83532;"}
  
    - *Best for: when all else fails*

## [1. R Documentation with `?`]{style="color:#BF396F;"}

Type `?function_name` in the console to open the help page for any function.

::: columns
::: {.column width="50%"}
```{r}
#| eval: false
?mean            #<1>
help(mean)       #<1>
```

1. These two commands do the same thing

 

Every help page has the same structure:

| Section | What it tells you |
|---|---|
| **Description** | What the function does |
| **Usage** | The function signature |
| **Arguments** | What inputs it takes |
| **Value** | What it returns |
| **Examples** | Copy-pasteable code |

:::

::: {.column width="50%"}
::: pink2
::: pink2-ttl
Pro tip: Jump straight to Examples
:::
::: pink2-cont
Scroll to the bottom of any help page — the **Examples** section is often the fastest way to understand how a function works.

```{r}
#| eval: false
example(mean)    #<1>
```

1. Run the examples directly
:::
:::

If `?` doesn't help, try searching across all installed packages:

```{r}
#| eval: false
??mean  #<1>      
```

1. Broader search of the function `mean()`
:::
:::

## [1. Demo: using `?` to solve a problem]{style="color:#BF396F;"}

::: pink2
::: pink2-cont
**Scenario:** You want to round the number `3.14159` to 2 decimal places. You know there's a `round()` function, but you're not sure how to use it.
:::
:::

::: columns
::: {.column width="50%"}
**Step 1:** Pull up the help page

```{r}
#| eval: false
?round
```


**Step 2:** Read the **Usage** and **Arguments** sections

- Arguments and Examples are very helpful sections!

```r
round(x, digits = 0)
```

- `x` — the number to round
- `digits` — how many decimal places (default is 0!)
:::

::: {.column width="50%"}
**Step 3:** Try it out

```{r}
round(3.14159)          #<1>
```

1. Default: 0 decimal places

```{r}
round(3.14159, digits = 2)   #<1> 
```

1. Now set `digits` to 2
:::
:::


## [2. Google and Stack Overflow]{style="color:#3070BF;"}

::: columns
::: {.column width="60%"}
::: blue
::: blue-cont
**Tips for a good search:**

- Always include **"in R"** or **"R programming"** in your query
- Copy and paste the **exact error message** into Google
- Add the **function or package name** if relevant
:::
:::

**Example:** Getting `NA` from `mean()`

> Google: `"mean function returning NA in R"`

This leads to answers like [this Stack Overflow thread](https://stackoverflow.com/questions/25297705/mean-na-rm-true-still-returns-na) or [this Posit Community post](https://forum.posit.co/t/why-r-return-na-for-all-my-functions/91751), where you'd learn about `na.rm = TRUE`.

```{r}
x <- c(1, 2, NA, 4)
mean(x)             #<1>
mean(x, na.rm = TRUE)  #<2>
```

1. Returns `NA` — not what we wanted!
2. Returns `2.333...` — ignores the missing value
:::

::: {.column width="5%"}
:::
::: {.column width="35%"}
::: blue
::: blue-ttl
Reading Stack Overflow answers
:::
::: blue-cont
- Check the **accepted answer** (green checkmark) first
- Look at the **vote count**: highly upvoted answers are usually reliable
- Read the **comments** too: they often clarify edge cases
- Check the **date**: R and its packages change over time
  - Ex: `mean( df$SleepDuration[df$Gender == 1] )` is a base R coding
:::
:::
:::
:::

## [2. Demo: Google and Stack Overflow]{style="color:#3070BF;"}

::: blue
::: blue-cont
**Scenario:** You try to combine two pieces of text in R and get an unexpected error.

```{r}
#| error: true
"Hello" + "World"    #<1>
```

1. Seems reasonable... but `+` is only for math!
:::
:::

 

::: columns
::: {.column width="50%"}
**Step 1:** Copy the error and search Google:

> Error in "Hello" + "World" : non-numeric argument to binary operator in R

::: blue
::: blue-cont
Googling the **exact error message** in quotes is often faster than describing the problem in your own words.
:::
:::

:::

::: {.column width="50%"}
**Step 2:** [Stack Overflow explains that `+` is only for math](https://stackoverflow.com/questions/66027835/non-numeric-argument-to-binary-operator-error-in-r) and [Google's search AI explains string combining uses `paste()` instead](https://share.google/aimode/wh2ZEyA8sJE9rogP8).

**Step 3:** Apply what you found

```{r}
paste("Hello", "World")   #<1>
paste0("Hello", "World")  #<2>
```

1. Combines with a space in between
2. Combines with no space

:::
:::

## [3. AI Tools (ChatGPT, Claude, etc.)]{style="color:#459B99;"}

::: columns
::: {.column width="50%"}
**When AI helps most:**

- Explaining what an error message means
- Getting starter code for something new
- Understanding what a function does

**Example prompts:**

> *"Why is the mean function in R giving me an NA?"*

> *"Can you explain what `na.rm = TRUE` does in R?"*

[See an example ChatGPT conversation →](https://chatgpt.com/share/66f8a2b9-9918-8006-88af-0dc7b64bb86a)
:::

::: {.column width="50%"}
**Watch out for:**

- AI often gives you **more code than you need**
- AI can be **confidently wrong** — always test the code it gives you
- AI may use **functions we haven't covered yet** — stick to tools from class notes

::: pink
::: pink-ttl
Course policy on AI
:::
::: pink-cont
If you cannot trace code back to the class notes, do **not** use it in assignments. You should be able to **explain every line** of your submitted code. See the [syllabus](https://nwakim.github.io/F24_EPI_525/syllabus.html#plagiarism-and-attribution) for details.
:::
:::
:::
:::

## [3. Demo: Using AI effectively]{style="color:#459B99;"}

::: green
::: green-cont
**Scenario:** You want to find the largest value in a vector but have no idea what function to use.
:::
:::

::: columns
::: {.column width="50%"}
**A vague prompt (less useful):**

> *"how do i get the biggest number"*

AI may give a long answer with many options you don't need yet.

**A better prompt:**

> *"In R, I have a vector of numbers: `c(4, 7, 2, 9, 1)`. What is the simplest function to find the maximum value?"*

AI will likely point you to `max()`.
:::

::: {.column width="50%"}
**Step 2:** Test it yourself — don't just trust it!

```{r}
scores <- c(4, 7, 2, 9, 1)
max(scores) #<1>
```

1. Always run the code to confirm it works

**Step 3:** Verify you understand it

```{r}
#| eval: false
?max  #<1>
```

1. Check the docs to confirm what it does

::: green
::: green-cont
AI gives you a **starting point**, not a final answer. Make sure you can explain every line before using it in your work.
:::
:::
:::
:::

## [4. Writing a Minimal Reproducible Example]{style="color:#C83532;"}

When asking for help — from a classmate, on Stack Overflow, or from AI — a **reprex** makes it much easier for others to help you.

A good reprex has three parts:

::: columns
::: {.column width="33%"}
**📦 Packages**

List every package you're using

```{r}
#| eval: false
library(tidyverse) #<1>
```

1. Include all packages needed to run your code
:::

::: {.column width="33%"}
**📋 Data**

Use a small built-in dataset, or `dput()` to share your own

```{r}
#| eval: false
head(mtcars)       #<1>

dput(head(my_data)) #<2>
```

1. Built-in datasets work great
2. Share a snippet of your own data
:::

::: {.column width="33%"}
**💥 The problem**

The exact code that produces the error

```{r}
#| error: true
mean(c("a", "b", "c")) #<1>
```

1. Paste the code that isn't working
:::
:::

::: pink2
::: pink2-cont
Focus on making your example **self-contained and minimal** — only include what's needed to show the problem.
:::
:::

## [4. Demo: Writing a reprex]{style="color:#C83532;"}

::: red
::: red-cont
**Scenario:** You're getting an unexpected result and want to ask your instructor or a classmate for help.
:::
:::

::: columns
::: {.column width="50%"}
**❌ A bad help request:**

> *"My code isn't working, can you help?"*

Nobody can debug this — there's no code, no error, no data.

**✅ A good help request:**

> *"I'm trying to calculate the mean of my numbers but keep getting NA. Here's my code:"*

```{r}
library(tidyverse)      #<1>

ages <- c(23, 31, NA, 45, 28) #<2>

mean(ages)              #<3>
```

1. Packages included
2. Small, simple data included
3. The exact code producing the problem
:::

::: {.column width="50%"}
**Why this works:**

- ✅ Anyone can copy-paste and run it immediately
- ✅ The data is right there — no guessing
- ✅ The expected vs. actual behavior is clearly stated

**The answer someone would give:**

```{r}
mean(ages, na.rm = TRUE)   #<1>
```

1. `na.rm = TRUE` tells R to ignore missing values

::: red
::: red-cont
Tell people: what you *wanted* to happen, what *actually* happened, and paste the full error message.
:::
:::
:::
:::

## A troubleshooting workflow

When you get stuck, try these steps **in order:**

1. **Read the error message carefully** — R is often telling you exactly what's wrong
2. **Check for typos** — a missing comma, parenthesis, or quote causes most beginner errors
3. **Use `?`** to check the function's expected arguments
4. **Google the error** — paste it in with "in R" and see what comes up
5. **Ask AI** — describe your problem clearly; paste in your code and error
6. **Ask a human** — office hours, a classmate, or a forum with a reprex

::: blue
::: blue-cont
Even experienced programmers spend a lot of time debugging. Getting better at reading error messages is one of the most valuable skills you'll build in this course.
:::
:::

## Summary

| Method | Best for | Try it when... |
|---|---|---|
| `?function_name` | Understanding arguments | You know the function, just not how to use it |
| Google / Stack Overflow | Specific errors | You have an error message to search |
| AI tools | Explanations, starter code | You want something explained or scaffolded |
| Office hours / classmates | Stuck after all else fails | You've tried the above and are still lost |

::: pink
::: pink-cont
Whatever help you use — you must be able to **read, understand, and explain** every line of code you submit.
:::
:::



## [2. Google and Stack Overflow]{style="color:#3070BF;"}

::: columns
::: {.column width="55%"}
**Tips for a good search:**

- Always include **"in R"** or **"R programming"** in your query
- Copy and paste the **exact error message** into Google
- Add the **function or package name** if relevant

**Example:** Getting `NA` from `mean()`

> Google: `"mean function returning NA in R"`

This leads to answers like [this Stack Overflow thread](https://stackoverflow.com/questions/25297705/mean-na-rm-true-still-returns-na), where you'd learn about `na.rm = TRUE`.

```{r}
x <- c(1, 2, NA, 4)
mean(x)             # Returns NA
mean(x, na.rm = TRUE)  # Returns 2.333...
```
:::

::: {.column width="45%"}
::: {.callout-tip}
## Reading Stack Overflow answers
- Check the **accepted answer** (green checkmark) first
- Look at the **vote count** — highly upvoted answers are usually reliable
- Read the **comments** too — they often clarify edge cases
- Check the **date** — R and its packages change over time
:::
:::
:::

## 2. Demo: Google and Stack Overflow

**Scenario:** You try to combine two pieces of text in R and get an unexpected error.

::: columns
::: {.column width="50%"}
**The problem:**

```{r}
#| error: true
"Hello" + "World"    # This seems reasonable... but fails!
```

**Step 1:** Copy the error message and search Google:

> `Error in "Hello" + "World" : non-numeric argument to binary operator in R`

**Step 2:** A Stack Overflow result explains that `+` is for math — string combining uses `paste()` instead.
:::

::: {.column width="50%"}
**Step 3:** Apply what you found

```{r}
paste("Hello", "World")          # with a space
paste0("Hello", "World")         # no space
```

::: {.callout-tip}
### Search tip
The phrase **"non-numeric argument to binary operator"** is a very common R error. Googling the exact error message in quotes is often faster than describing it in your own words.
:::
:::
:::

## [3. AI Tools (ChatGPT, Claude, etc.)]{style="color:#459B99;"}

AI can be a useful coding assistant — but you need to use it wisely.

::: columns
::: {.column width="50%"}
**When AI helps most:**

- Explaining what an error message means
- Getting starter code for something new
- Understanding what a function does

**Example prompts:**
> "Why is the mean function in R giving me an NA?"

> "Can you explain what `na.rm = TRUE` does in R?"

[See an example ChatGPT conversation →](https://chatgpt.com/share/66f8a2b9-9918-8006-88af-0dc7b64bb86a)
:::

::: {.column width="50%"}
**Watch out for:**

- AI often gives you **more code than you need** — focus on extracting the piece that answers your question
- AI can be **confidently wrong** — always test the code it gives you
- AI may use **functions or packages we haven't covered** — stick to tools from class notes

::: {.callout-warning}
## Course policy on AI
If you cannot trace code back to the class notes, do **not** use it in assignments. You should be able to **explain every line** of your submitted code. See the [syllabus](https://nwakim.github.io/F24_EPI_525/syllabus.html#plagiarism-and-attribution) for details.
:::
:::
:::

## [3. Demo: Using AI effectively]{style="color:#459B99;"}

**Scenario:** You want to find the largest value in a vector but you have no idea what function to use.

::: columns
::: {.column width="50%"}
**A vague prompt (less useful):**
> *"how do i get the biggest number"*

AI may give you a long answer with many options and extra explanation you don't need yet.

**A better prompt:**
> *"In R, I have a vector of numbers: `c(4, 7, 2, 9, 1)`. What is the simplest function to find the maximum value?"*

AI response will likely point you to `max()`.
:::

::: {.column width="50%"}
**Step 2:** Test it yourself — don't just trust it!

```{r}
scores <- c(4, 7, 2, 9, 1)
max(scores)
```

**Step 3:** Verify you understand it

```{r}
#| eval: false
?max    # check the docs to confirm what it does
```

::: {.callout-warning}
## AI gives you a starting point, not a final answer
Always run the code yourself and make sure you can explain what it does before using it in your work.
:::
:::
:::

## [4. Writing a Minimal Reproducible Example]{style="color:#C83532;"}

When asking for help — from a classmate, on Stack Overflow, or from AI — a **reprex** (reproducible example) makes it much easier for others to help you.

A good reprex has three parts:

::: columns
::: {.column width="33%"}
**📦 Packages**

List every package you're using

```{r}
#| eval: false
library(tidyverse)
```
:::

::: {.column width="33%"}
**📋 Data**

Use a small built-in dataset, or `dput()` to share your own

```{r}
#| eval: false
# Use a built-in dataset
head(mtcars)

# Or share a snippet of yours
dput(head(my_data))
```
:::

::: {.column width="33%"}
**💥 The problem**

The code that produces the error

```{r}
#| error: true
mean(c("a", "b", "c"))
```
:::
:::

::: {.callout-tip}
## The `reprex` package
The `reprex` package formats your example for easy pasting into Stack Overflow or GitHub. We'll keep it simple for now — just focus on making your example **self-contained and minimal**.
:::

## [4. Demo: Writing a reprex]{style="color:#C83532;"}

**Scenario:** You're getting an unexpected result and want to ask your instructor or a classmate for help.

::: columns
::: {.column width="50%"}
**❌ A bad help request:**

> *"My code isn't working, can you help?"*

Nobody can debug this — there's no code, no error, no data.

**✅ A good help request (reprex):**

> *"Hi! I'm trying to calculate the mean of my numbers but keep getting NA. Here's my code:"*

```{r}
# Packages
library(tidyverse)

# Data
ages <- c(23, 31, NA, 45, 28)

# Problem
mean(ages)     # I expected a number but got NA
```
:::

::: {.column width="50%"}
**Why this works:**

- ✅ Anyone can copy-paste and run it immediately
- ✅ The data is right there — no guessing
- ✅ The expected vs. actual behavior is clearly stated

**The answer someone would give:**

```{r}
mean(ages, na.rm = TRUE)   # ignore the missing value
```

::: {.callout-tip}
## When in doubt, include more context
Tell people: what you *wanted* to happen, what *actually* happened, and paste the full error message.
:::
:::
:::

## A troubleshooting workflow

When you get stuck, try these steps **in order:**

1. **Read the error message carefully** — R is often telling you exactly what's wrong
2. **Check for typos** — a missing comma, parenthesis, or quote causes most beginner errors
3. **Use `?`** to check the function's expected arguments
4. **Google the error** — paste it in with "in R" and see what comes up
5. **Ask AI** — describe your problem clearly; paste in your code and error
6. **Ask a human** — office hours, a classmate, or a forum with a reprex

::: blue
::: blue-cont
Even experienced programmers spend a lot of time debugging. Getting better at reading error messages is one of the most valuable skills you'll build in this course.
:::
:::

## Summary

| Method | Best for | Try it when... |
|---|---|---|
| `?function_name` | Understanding arguments | You know the function, just not how to use it |
| Google / Stack Overflow | Specific errors | You have an error message to search |
| AI tools | Explanations, starter code | You want something explained or scaffolded |
| Office hours / classmates | Stuck after all else fails | You've tried the above and are still lost |

::: pink
::: pink-cont
Whatever help you use — you must be able to **read, understand, and explain** every line of code you submit.
:::
:::


