---
title: "Functions"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Functions
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: below
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
```

## Using functions

-   `mean()` is an example of a function
-   functions have "arguments" that can be specified within the `()`
-   `?mean` in console will show help file for `mean()`

::: columns
::: {.column width="50%"}
Function <font color="#BF396F">arguments specified</font> by name:

```{r}
mean(x = 1:4)
```

```{r}
seq(from = 1, to = 12, by = 3)
seq(by = 3, to = 12, from = 1)
```
:::

::: {.column width="50%"}
Function <font color="#BF396F">arguments not specified</font>, but listed in order:

```{r}
mean(1:4)
```

```{r}
seq(1, 12, 3)
```
:::
:::

## Now let's use some functions for summary statistics

-   We will calculate the mean for `c`
-   Let's also calculate the standard deviation for `c`
    -   Recall, our function is `sd()`
    -   Use `?sd` in the console to identify the arguments for `c`

 

-   If you have more time, you can try to calculate the median and IQR for `c`

```{r}
#| echo: false
countdown::countdown(3)
```
