---
title: "Data transformation: Columns"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: Columns"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(readxl)
library(gt)
library(rstatix)
```

```{r}
#| echo: false
# library(readxl)
# note we are saving the data from the Excel file in our environment 
# with the object name hrs_data
hrs_data2 <- read_excel(here("./data/hrs_data_narch.xlsx"))
hrs_data <- read_excel(here("./data/hrs_data.xlsx"))
hrs_data <- clean_names(hrs_data) %>%
  relocate(race_original, female, .after = degree)#
```

## Data transformations on columns

- There are many functions that can be used to transform columns
- We can use columns to create new variables, subset features, and reorganize a dataset
- We will cover the most column functions:
  - `mutate()`
  - `select()`
  - `rename()`
  - `relocate()`

# `mutate()`

![[Artwork by \@allison_horst](https://allisonhorst.com/everything-else){target="_blank"}](../img_slides/horst_mutate.jpg){fig-align="center"}


## `mutate()`: transform and create variables

- `mutate()` allows us to transform existing columns and/or add new columns to the dataset

 

- Think of this like adding a formula in Excel to calculate the value of a new column based on previous columns

 

- `mutate()` is useful for:
  - Converting the units of one column to new units (e.g., months to years)
  - Performing arithmetic using multiple columns
  - Changing variable types or recoding categories

## Example: calculating a new variable

- Let's convert `age_months` into years by dividing it by 12
- We will name this new variable `age_yrs`

```{r}
hrs_data |> 
  mutate(
    age_yrs = age_months / 12 # <1>
  )
```

1.  **Notice we use a single `=` inside mutate**, not `==` or `<-`. This sets our new column name equal to the mathematical operation.

## Example: calculating multiple columns and saving

- We can create multiple new columns within a single `mutate()` function by separating them with a comma
- Let's save our work to a new object called `hrs_data2` to avoid overwriting our original dataset

```{r}
hrs_data2 <- hrs_data |> # <1>
  mutate(
    age_yrs = age_months / 12, # <2>
    age_decades = age_yrs / 10 # <3>
  )
```

1.  We use the assignment operator `<-` to save the output into a brand new data frame called `hrs_data2`.
2.  First, we create the `age_yrs` column.
3.  Second, we create `age_decades` using the `age_yrs` variable we *just* defined in the step right above it!

- Let's take a look at the new data frame `hrs_data2` to see our new columns

# `select()`

## Subsetting columns using `select()`

- The `select()` function allows us to isolate specific variables or columns from our dataset
- Why use `select()`?
  - Drops unneeded information to **maximize processing speed and visual clarity**
  - Isolates specific blocks of data for focused modeling or visualization

## Example: selecting specific columns

- Let's keep only three variables from our dataset: `age`, `smoke2now`, and `diab`
  - Column names do NOT need quotes in `select()`

```{r}
hrs_data2 |> 
  select(
    age, smoke2now, diab # <1>
  )

```

1.  We provide the exact column names we want to keep, separated by commas. All other columns will be dropped from the output.

## Example: removing specific columns

- If we want to keep everything **except** a few specified variables, we use a minus sign (`-`)

```{r}
hrs_data2 |> 
  select(
    -drink, -lung_dis # <1>
  )

```

1.  Placing `-` in front of `drink` and `lung_dis` explicitly drops them, while retaining every other original variable.

## Example: selecting consecutive columns

- We can select a continuous block of columns using the `:` operator

```{r}
hrs_data2 |> 
  select(
    age:drink # <1>
  )
```

1.  This keeps `age`, `drink`, and every single column positioned between them in the data frame. **Make sure you know your column order before using this!**

## Distinguishing `filter()` vs `select()`

# `rename()`

## Renaming with `rename()`

-   You can rename columns with `select()` but it's a bit easier to do this with the `rename()` function.

-   You just need to remember `new_name = old_name` ordering.

```{r}
hrs_data3 <- hrs_data2 |> 
  rename(
    id = hhid,
    race = race_original,
    sex_bin = female
    )
```

# `relocate()`

## Reordering columns using `relocate()`

- `relocate()` moves columns to a different horizontal position in your dataset
- By default, variables sent into `relocate()` move directly to the **very front** of the dataset

```{r}
#| eval: false
hrs_data3 |> relocate(lung_dis, arthritis) # <1>
hrs_data3 |> relocate(lung_dis, .after = birthdate)       # <2>
hrs_data3 |> relocate(lung_dis, .before = race)        # <3>
```

1.  Moves `lung_dis` and `arthritis` to the very front of the dataset, before all other columns.
2.  Moves `income` to place it directly behind the `record` column.
3.  Moves `income` to place it right before the `age` column.

- Let's take a look at the output!

## Wrap-up

- We learned three functions that allow us to transform columns in our dataset:
  - `mutate()`
    - Transform existing columns and/or create new columns
  - `select()`
    - Subset specific columns to keep or drop
  - `rename()`
    - Rename columns with new names
  - `relocate()`
    - Reorder columns to a different spot in the dataset
